Spectraplakin cooperates with noncentrosomal microtubule regulators to orient dendritic microtubules in Drosophila

Version 1

Abstract

Abstract

The differential microtubule organization in axons and dendrites underlies neurite-specific transport and morphogenetic processes like developmental neurite pruning, but how it arises is incompletely understood. While axonal "plus end-out" microtubules just grow from the soma, the origin of the dendritic "plus end-in" microtubules is more difficult to explain. A transient microtubule organizing center (MTOC) was recently detected at dendrite tips in C. elegans, but the players and generalizability are unclear. In a survey for factors involved in dendrite pruning of Drosophila sensory neurons, we found the spectraplakin Short stop (Shot), an actin-microtubule crosslinker known to bundle axonal microtubules. In dendrites, however, Shot is required for proper "plus end-in" microtubule organization. During early dendrite development, Shot localizes to dendrite tips in an actin-dependent manner, providing a potential mechanism for dendrite tip recognition. Our data show that Shot cooperates with a functional MTOC containing the small GTPase Rab11. Localized MTOC activity may commonly underlie dendritic microtubule organization.

Introduction

Microtubules are long polar tube-like polymers composed of α/β-tubulin dimers that provide mechanical stability to cells and act as tracks for motor-based transport. Tubulin dimers are preferentially added, and also lost, at microtubule plus ends, whereas microtubule minus ends are often anchored at microtubule organizing centers (MTOCs), structures of diverse composition that nucleate and anchor microtubules. In dividing cells, centrosomes are the main MTOCs, but many non-dividing or differentiated cells have less well characterized non-centrosomal (nc)MTOCs (Akhmanova & Kapitein, 2022).

Given their intrinsic polarity, microtubules predefine neuronal polarity. In axons, all microtubules are oriented with their plus ends towards the distal synapses, thus enabling axon-specific transport via kinesin-1. In contrast, a large fraction of dendritic microtubules are oriented with their plus ends towards the soma ("plus end-in"), enabling dendrite-specific transport via other motors including dynein. This orientation is particularly evident in invertebrates, where dendritic microtubules in primary dendrite branches almost uniformly adopt the plus end-in orientation. Despite the recognized importance of microtubule organization for neuronal function, it is still poorly understood how neuronal microtubule organization is achieved and maintained.

The differential organization of neuronal microtubules also underlies important developmental processes. The peripheral nociceptive class IV dendritic arborization (c4da) neurons of Drosophila larvae prune their long and branched sensory dendrites at the onset of metamorphosis. Pruning involves severing of proximal dendrite regions and subsequent degeneration of distal fragments (Rumpf et al, 2017; Furusawa & Emoto, 2021). The presumptive severing sites are marked by the local loss of microtubules, which predisposes them for mechanical tearing (Krämer et al, 2023). This local microtubule disassembly is initiated by an increase in microtubule dynamics at the onset of the pupal stage (Herzmann et al, 2017). Due to their uniform plus end- in orientation, dendritic microtubules can then shrink from their plus ends in a coordinated fashion, leading to their local loss in proximal dendrite regions (Herzmann et al, 2018; Rumpf et al, 2019). In support of this model, conditions that cause misorientation of dendritic microtubules bring about pruning defects. For example, pruning defects are associated with mutations in the kinesin-2 motor, which guides newly polymerized microtubules along preexisting ones, thus ensuring the identical orientation (Mattie et al, 2010; Herzmann et al, 2018). Microtubule orientation and pruning defects are also caused by loss of the minus end binding protein patronin (Wang et al, 2019), the microtubule polymerase/nucleation factor Minispindles (Msps)/XMAP-215 (Tang et al, 2020) and the serine/threonine phosphatase PP2A (Rui et al, 2020; Wolterhoff et al, 2020), but the underlying mechanisms are not always clear. An additional mechanism for dendritic microtubule orientation was described in C. elegans PVD neurons, where the localization of dendritic MTOCs is important for the establishment of uniform plus end-in organization. Here, a prominent ncMTOC resides at the tips of growing dendrites, such that microtubules can only grow towards the soma (Liang et al, 2020; Harterink et al, 2018). This ncMTOC was shown to reside on Rab11-positive recycling endosomes, and Rab11 was important for microtubule orientation (Liang et al, 2020).

Spectraplakins are a family of actin/microtubule cytoskeletal crosslinker proteins with important roles in microtubule organization in differentiated cells (Sanchez et al, 2021; Sun et al, 2019; Voelzmann et al, 2017). Anchored to the actin cortex via an N-terminal actin binding domain, they can recruit microtubules either directly via a GAS2-related domain, or via binding sites for end binding proteins (Noordstra et al, 2016; Ning et al, 2016; Nashchekin et al, 2016) (Slep et al, 2005). The Drosophila spectraplakin Short stop (Shot) is known to bundle and stabilize microtubules in axons (Alves-Silva et al, 2012, Okenve-Ramos et al, 2024).

Here, we describe an important role for Shot in dendrites. Loss of Shot causes pruning defects caused by mixed orientation of dendritic microtubules. During early dendrite growth stages, Shot localizes to dendrite tips in an actin-dependent manner where it recruits factors cooperating with an early-acting, Rab11-dependent MTOC. Our data provide a mechanism for MTOC tip recruitment for the establishment of oriented microtubule arrays in dendrites.

Results

Shot is required for c4da neuron dendrite pruning

C4da neurons have long and branched dendrites at the third instar larval (L3) stage (Fig. 1 A) that are pruned at the onset of the pupal stage, such that by 16 hours after puparium formation (h APF), they have lost all their dendrites (Fig. 1 A’, G - I). In order to identify new regulators of c4da neuron dendrite pruning, we expressed candidate RNAis in c4da neurons under pickpocket-GAL4 and assessed their effects on dendrite morphology and pruning. This approach identified a dsRNA line targeting Shot, which encodes the sole Drosophila member of the spectraplakins, a family of large adaptor proteins with binding sites for actin and microtubules. C4da neurons expressing shot dsRNA did not display major dendritic phenotypes at the larval stage (Fig. 1 B), even though their dendritic field coverage seemed somewhat reduced compared to control neurons. In contrast, dendrites could still be seen attached to the soma at 16 h APF in almost half the c4da neurons expressing shot dsRNA (Fig. 1 B’, G - I), indicative of pruning defects. In order to confirm this result, we used Mosaic Analysis with a Repressible Cell Marker (MARCM) to generate c4da neuron clones homozygous for the loss-of-function allele shot3 (Kolodziej et al, 1995). shot3 mutant c4da neurons had shorter dendrites at the L3 stage compared to control MARCM c4da neuron clones (Fig. 1 C, D). At 16 h APF, most shot3 mutant neurons still had dendrites attached to the soma (Fig. 1 C’ D’, G - I). We also generated an sgRNA line targeting a common exon of all Shot splice isoforms and found that it caused significant c4da neuron dendrite pruning defects when expressed in c4da neurons in a sensitized shot3/+ heterozygous background (Fig. 1 E’ -F’, G - I).

Shot is required for c4da neuron developmental dendrite pruning.

A – F’ C4da neurons of the indicated genotypes were imaged at the third instar larval stage (A - F) and at 16 hours after puparium formation (h APF) (A’ - F’). A, A’ Control c4da neurons expressing Orco dsRNA under the control of ppk-GAL4. B, B’ C4da neurons expressing Shot dsRNA under the control of ppk-GAL4. C, C’ Control c4da neurons labeled by MARCM. D, D’ shot3 mutant c4da neurons labeled by MARCM. E, E’ Control c4da neurons expressing Cas9P2 under ppk-GAL4 in a shot3 heterozygous background. F, F’ C4da neurons coexpressing Cas9P2 and Shot sgRNA under the control of ppk-GAL4 in a shot3 heterozygous background. G Phenotypic penetrance of pruning defects in A’ - F’. Sample sizes are indicated above the graph. *** P<0.001, **** P<0.0001, two-tailed Fisher’s exact test. H Average number of primary and secondary dendrites attached to the soma at 16 h APF in samples A’ - F’. Values are mean +/- s. e. m., **** P<0.0001, Mann-Whitney U test. I Total length of remaining dendrites at 16 h APF in samples A’ - F’. Values are mean +/- s. e. m., **** P<0.0001, Mann-Whitney U test. Scale bars in A and A’ are 50 µm.

Dendrite pruning is the first step of c4da neuron metamorphic remodeling and is followed by the regeneration of new dendrites adapted to the adult stage. We noted that c4da neurons lacking Shot also failed to initiate dendrite regrowth at 72 h APF, when control c4da neurons had regrown long and branched dendrites (Figure S1). We conclude that Shot is broadly required for c4da neuron dendrite remodeling.

Shot is required for uniform dendritic microtubule orientation

We and others have previously shown that c4da neuron dendrite pruning depends on local microtubule disassembly in proximal dendrites (Herzmann et al, 2017). As Shot is a microtubule regulator, we assessed the state of dendritic microtubules by immunofluorescence against the microtubule-associated protein futsch/MAP1B. At 5 h APF, continuous futsch staining could often still be detected in proximal dendrites of shot knockdown neurons, while control neurons had large gaps their futsch signal in these regions (Fig. S2). The uniform plus end-in orientation of dendritic microtubules is particularly important for dendrite pruning as microtubules must shrink away from the cell body in a coordinated fashion (Herzmann et al, 2018; Rumpf et al, 2019, Wang et al, 2019). We therefore asked whether loss of Shot affects dendritic microtubule orientation by using EB1::GFP to monitor the microtubule growth direction. In primary dendrites of third instar control c4da neurons, EB1::GFP comets (the plus ends of growing microtubules) moved exclusively retrogradely towards the soma (Fig. 2 A, C).

Shot is required for uniform plus end-in orientation of dendritic microtubules.

A, B EB1::GFP comets were imaged in primary dendrites of third instar larval c4da neurons, and comet movement was depicted in kymographs. Direction of the soma and time are indicated. A Upper panel: EB1::GFP comets in control neuron expressing Orco dsRNA; lower panel: EB1::GFP comets in neuron expressing Shot dsRNA. B Upper panel: EB1::GFP comets in control c4da neuron MARCM clone; lower panel: EB1::GFP comets in shot3mutant c4da neuron MARCM clone. C Percentage of anterogradely moving comets in panels A, B. N is indicated above the graph. ** P<0.01, **** P<0.0001, two-tailed Fisher’s exact test. D EB1::GFP comet speed in panels A, B. **** P<0.0001, Wilcoxon’s test. The scale bar in A is 5 µm.

Upon shot knockdown, however, a significant proportion of EB1::GFP comets could be seen moving anterogradely (Fig. 2 A, C). In dendrites of shot3 c4da neuron MARCM clones, we observed an even stronger increase in anterogradely moving comets in comparison to control MARCM clones (Fig. 2 B, C). In support of a broad role for Shot in dendritic microtubule organization, EB1 comets in shot3 mutant c4da neurons also moved at an increased speed compared to controls (Fig. 2 D). We conclude that Shot is required for the correct orientation of dendritic microtubules, and that this function likely underlies its role in pruning.

Actin and microtubule binding are crucial for Shot function in dendrites

In order to better understand the function of Shot in dendritic pruning and microtubule orientation, we next asked which domains of Shot are crucial for these functions. Of particular interest were the domains of Shot that bind actin and microtubules. Shot has two Calponin Homology (CH) domains at the N-terminus that mediate binding to specific actin structures, and a microtubule-binding GAS2 domain in its C-terminal region (Fig. 3 A). Shot also possesses functionally important binding sites for end binding proteins such as Patronin and EB1. We next expressed full length or truncated Shot variants in shot3 mutant c4da neurons using the GAL4/UAS system and asked whether they could rescue the pruning defects of the mutant. A full length Shot UAS transgene (ShotFL) fully rescued the pruning defects of shot3 mutant neurons (Fig. 3 B, C, F, G). In contrast, ShotβCH1::GFP, a Shot variant lacking the first CH domain that has decreased affinity for actin (Lee & Kolodziej, 2002), afforded only a partial rescue (Fig. 3 D, F, G). Interestingly, overexpression of a C-terminal Shot fragment containing the microtubule binding domain and adjacent regions, but not the CH or spectrin repeat domains (UAS-Cterm::YFP), also fully rescued the pruning defects of mutant cells (Fig. 3 E - G). In further support of non-functionality, overexpression of ShotβCH1 caused strong pruning defects (Fig. S3).

Shot domains important for pruning and microtubule organization.

A UAS-Shot constructs. CH, Calponin homology. B - E Ability of the indicated Shot UAS constructs to rescue the pruning defects of shot3 mutant c4da neurons at 16 h APF. B shot3 mutant c4da neuron labeled by MARCM. C shot3 mutant c4da neuron expressing full-length Shot::GFP. D shot3mutant c4da neuron expressing Shot::GFP lacking the CH1 domain. E shot3 mutant c4da neuron expressing Shot C-term::YFP. F Penetrance of pruning defects. N is indicated above the graph. * P<0.05, *** P<0.001, **** P<0.0001, two-tailed Fisher’s exact test. G Number of primary and secondary dendrites attached to the cell body at the indicated timepoints. Values are mean +/- s. e. m., *** p<0.001, **** p<0.0001, Mann-Whitney U test. H – L Kin::lacZ localization in third instar c4da neuron MARCM clones. H Control c4da neuron. I shot3 mutant c4da neuron. J shot3mutant c4da neuron expressing full-length Shot. K shot3mutant c4da neuron expressing Shot lacking the CH1 domain. L shot3mutant c4da neuron expressing Shot C-term::YFP. Arrows in H – L mark dendritic kin::lacZ puncta, asterisks denote soma position. M Number of dendritic kin::lacZ puncta in third instar c4da neurons expressing the indicated Shot UAS constructs. N is depicted above the graph. Values are mean +/- s. e. m., * p<0.05, ** p<0.01, *** p<0.001, Wilcoxon test. Scale bars in B and H are 50 μm.

In order to assess the functionality of these Shot variants in dendritic microtubule organization, we made use of kinesin-lacZ (kin::lacZ), a fusion between the kinesin-1 motor domain and β-galacosidase. In control MARCM c4da neurons, kinesin-lacZ localizes exclusively to the soma and axon (Fig. 3 H), as the kinesin motor travels towards microtubule plus ends. In shot3 mutant c4da neurons, axonal kin::lacZ was reduced, and it could instead be found in distinct puncta in dendrites, indicating the presence of plus end-out microtubules in these dendrites (Fig. 3 I, M). Re-expression of full length Shot, but not of ShotβCH1::GFP, significantly rescued the number of dendritic kin::lacZ puncta (Fig. 3 J, K, M). As with the pruning defects, overexpression of Shot-Cterm::YFP restored dendritic exclusion of kin::lacZ (Fig. 3 L, M).

It had previously been suggested that Shot and its mammalian homologues can bundle and/or stabilize microtubules. In order to assess the effects of Shot loss on dendritic microtubule structure, we labeled microtubules in c4da neurons by expression of mcherry-tagged α-tubulin and used 2D stimulated emission depletion (STED) microscopy for visualization. In primary dendrites of control c4da neurons, mCherry- tagged microtubules appeared as several long filaments (Fig. S4). In dendrites of c4da neurons lacking Shot, fewer filamentous structures were visible and the staining was more punctate, possibly indicating that these neurons have fewer and/or damaged microtubules (Fig. S4). Thus, both microtubule and actin binding activities of Shot are important for correct dendritic microtubule orientation and pruning.

Shot cooperates with actin to promote microtubule orientation

The above domain analysis demonstrated that the actin binding ability of Shot is important for its function. Spectraplakins often bind to cortical actin structures underneath the plasma membrane. Consistently, we found that endogenously GFP- tagged Shot localized to the rim of the dendrite in third instar larval c4da neuron dendrites (Fig. 4 A, A’) in structured illumination microscopy (SIM) imaging. To directly probe for a role for actin in dendritic microtubule organization, we sought to disassemble actin filaments in vivo. The actin-severing enzyme Mical is a pruning factor that is induced at the onset of metamorphosis (Kirilly et al, 2009). We used the GAL4/UAS system to overexpress Mical prematurely in larval c4da neurons. In dendrites of c4da neurons overexpressing Mical, a significant fraction of EB1::GFP comets moved anterogradely, indicating mixed microtubule orientation (Fig. 4 B, C). Mical overexpression combined with Shot knockdown did not lead to an additive increase in the percentage of anterograde comets, indicating epistasis between the manipulations (Fig. 4 B, C). However, these anterograde comets moved significantly faster than their retrograde counterparts (Fig. 4 D). Furthermore, movement of the anterograde comets in this sample was often irregular, and comets frequently slowed down and then sped up again (Fig. 4 B). Such phenomena are often seen when microtubules are not attached to specific anchoring sites such that they can be moved by microtubule motors.

Shot links microtubule regulation to actin.

A, A’ Localization of Shot in a third instar primary c4da neuron dendrite. Endogenously tagged Shot::GFP was labeled by immunofluorescence and visualized by structured illumination microscopy. C4da neurons were labeled by expression of cytosolic tdtomato under ppk-GAL4. A Shot::GFP. A’ merge with tdtomato. B Effect of actin severing on EB1::GFP comet movement. Mical was overexpressed in c4da neurons with or without Shot knockdown, and EB1 comets were analyzed as in Fig. 2 B. Upper panel, kymograph of EB1 comet movement in c4da neuron overexpressing Mical. Lower panel, kymograph of EB1 comet movement upon Mical overexpression and shot knockdown. C Penetrance of anterograde comets in Fig. 4 B. N is given in the graph. * P<0.05, Fisher’s exact test. D Graph depicting the effects of Shot knockdown and Mical expression on EB1 comet speed. N is given in the graph. **** P<0.0001, Wilcoxon’s test. E Synergistic effects of actin severing and Shot knockdown on c4da neuron dendrite pruning. Penetrance of pruning defects at 16 h APF in c4da neurons overexpressing Mical, upon Shot knockdown, or both. N is given in the graph. * P<0.05, Fisher’s exact test. Scale bars in A and B are 5 μm.

Mical overexpression suppresses the pruning defects of ecdysone signaling mutants (Kirilly et al, 2009; Rode et al, 2018). However, we observed a synergistic enhancement of the pruning defects caused by Shot knockdown when we overexpressed Mical (Fig. 4 E). Thus, Shot’s role in dendrite pruning might in part be related to microtubule anchoring.

Shot is recruited to dendritic tips during early dendrite growth

Our analysis of Shot domains showed that the Shot CH1 domain is important for dendrite pruning and microtubule orientation (Fig. 3). This domain is often required to recruit Shot to specific subcellular locations (Nashchekin et al, 2024). To test whether the CH1 domain is required for localization to the dendrite membrane (Fig. 4 A), we visualized transgenic Shot or ShotΔCH1 in third instar c4da neurons using SIM. Similar to endogenously GFP-tagged Shot, transgenic full length Shot::GFP localized in puncta and sometimes streaks (especially around dendritic branchpoints) close to the plasma membrane (Fig. 5 A). ShotΔCH1::GFP localized in a similar pattern to full-length Shot (Fig. 5 B), indicating that the CH1 domain is unlikely to be required for Shot cortical localization in dendritic shafts.

Shot is recruited to tips of growing dendrites via its CH1 domain.

The indicated GFP-tagged Shot constructs and tdtomato were expressed in c4da neurons under ppk-GAL4 and visualized at the indicated developmental stages. A, B Dendritic shaft localization of transgenic Shot::GFP variants in third instar c4da neuron dendrites was visualized by immunofluorescence and structured illumination microscopy (SIM). A Localization of full-length Shot (ShotFL::GFP). B Localization of Shot lacking the CH1 domain (ShotΔCH1::GFP). C - H Localization of Shot variants in first instar c4da neurons. Images on the right show close-ups of dendrite tip (1) and soma regions (2). The asterisk in the dendrite close-up marks the position of the tip. C Localization of ShotFL::GFP. D ShotΔCH1::GFP. E Fluorescence intensity profiles of Shot::GFP and ShotΔCH1::GFP in the distal 10 μm of first instar dendrites. Intensity values were normalized to Shot::GFP intensity in the cell body. Solid lines indicate average, envelopes indicate S. D. (N=10). The table shows significance between genotypes for the indicated distances from the tip **** P<0.0001, *** P<0.001, n. s. not significant, Mann Whitney U test. F Dendrite tip occupancy of ShotFL::GFP and ShotΔCH1::GFP. ** P<0.01, Fisher’s exact test. G Effect of Mical overexpression on ShotFL::GFP localization. Images on the right show close-ups of dendrite tip (1) and soma regions (2). H Fluorescence intensity profiles of ShotFL::GFP in the distal 10 μm of first instar dendrites with or without Mical overexpression. Quantification was as in E. N=10, **** P<0.0001, *** P<0.001, ** P<0.01, n. s. not significant, Mann Whitney U test. Scale bars are 5 μm in A and 10 μm in C (larger image and close-up).

Since dendritic microtubule organization is set up early during development, we investigated the localization of transgenic Shot::GFP in c4da neurons at the first larval instar stage, when c4da neuron dendrites are growing very fast. At this stage, Shot::GFP was highly enriched at dendrite tips, but could only be detected at low levels in the soma and the more proximal dendritic shafts (Fig. 5 C, E). Shot tip enrichment seemed transient, as it was not observed in mature c4da neuron dendrites at the third instar (Fig. S5). In contrast to full length Shot, ShotβCH1::GFP showed significantly lower tip occupancy (Fig. 5 D, F) and often localized in dots that could be several micrometers away from the tips (Fig. 5 D, E). To independently test whether Shot tip localization was actin-dependent, we used UAS-Mical for enzymatic actin disruption. Mical overexpression did not abolish Shot at tips, but caused a significant broadening of the tip signal, again indicating disturbed Shot recruitment (Fig. 5 G, H). Thus, efficient dendrite tip recruitment of Shot during the early stages of c4da neuron dendrite growth requires the CH1 domain.

Rab11 is a component of a developmental dendritic MTOC

The fact that Shot variants lacking the CH1 domain cannot rescue the pruning defects of shot3 mutants suggested that dendrite tip localization of Shot was important for its function. The tips of growing dendrites have recently been linked to microtubule organization in C. elegans PVD sensory neurons. Here, tip-localized Rab11-positive vesicles recruit a ncMTOC (Liang et al, 2020; Harterink et al, 2018). The composition and localization of dendritic MTOCs in Drosophila sensory (c4da and c1da) neurons are debated. Microtubules have been observed originating in dendritic shafts, at branchpoints and at tips, and both Golgi outposts and Rab5 endosomes have been implicated as sites harbouring MTOCs (Ori-McKenney et al, 2012; Nguyen et al, 2014; Mukherjee et al, 2020; Yalgin et al, 2015; Weiner et al, 2020). Most studies in c4da neurons have focused on mature dendrites in third instar neurons. We therefore asked whether the same factors and/or a similar mechanism as in PVD neurons also act in early developing c4da neurons.

Rab11 is required for c4da neuron dendrite pruning (Krämer et al, 2019; Lin et al, 2020). Given the strong dependence of pruning on microtubule organization, it is interesting to speculate that it might perform a microtubule-related function in early c4da neuron dendrites as well. We first tested whether Rab11 colocalizes with the bona fide MTOC component γ-tubulin. To this end, we coexpressed Rab11::mCherry with GFP-tagged γ-tubulin in c4da neurons. At the first instar, a subset of Rab11::mCherry puncta - likely vesicles - could be seen overlapping with γ- tubulin::GFP in discrete dendritic puncta that could also be found at dendrite tips (Fig. 6 A). To test whether such puncta could nucleate microtubules, we observed Rab11::mCherry together with EB1::GFP. In first instar neurons, EB1::GFP comets could be observed originating from dendritic Rab11::mCherry puncta (6/9 neurons, 10/58 comets) (Fig. 6 B, C). These comets were mostly seen in dendritic shafts or at branchpoints, likely because we could only capture a subset of EB1::GFP comets in our experiments. No EB1::GFP comets could be observed from Rab11::mCherry puncta at the third instar (0/9 neurons, 27 comets) (Fig. 6 C). To test for a functional role for Rab11 in dendritic microtubule organization, we knocked it down and visualized growing microtubules at the third instar stage. Rab11 knockdown caused a significant fraction of EB1 comets (16 %) to move anterogradely, and overall comet speed was increased (Fig. 6 D - F). In contrast, a dsRNA targeting Rab5 did not cause a significant increase in anterograde comets in c4da neurons, and comet speed was normal (Fig. 6 D - F).

Evidence that Rab11 is part of a developmentally regulated dendritic microtubule organizing center.

A Rab11 colocalizes with the MTOC marker γ-tubulin in growing c4da neuron dendrites. Rab11::mCherry and γ-tubulin23C::GFP were coexpressed under ppk-GAL4 and visualized at the first instar. Arrows show Rab11/ γ-tubulin-positive puncta at dendrite tips, the asterisk denotes a double-positive dot at a branchpoint. B Microtubules can be nucleated at Rab11 puncta in dendrites. Rab11::mCherry and EB1::GFP were coexpressed in c4da neurons, and EB1 comets were visualized at the first instar. Example kymographs show Rab11::mCherry (magenta) and EB1::GFP (green). C Percentage of EB1 comets arising from dendritic Rab11 puncta at the first and third instars. * P<0.05, Fisher’s exact test. D Effect of Rab11 knockdown on dendritic microtubule orientation. Kymographs show EB1::GFP movement in third instar c4da neuron dendrites. Upper panel, control c4da neuron expressing Orco dsRNA; middle panel, c4da neuron expressing Rab11 dsRNA; lower panel, c4da neuron expressing Rab5 dsRNA. E Penetrance of anterograde comets in D. ** P<0.01, two-tailed Fisher’s exact’s test. F Speed of EB1::GFP comets in D. **** P<0.0001, Wilcoxon’s test. G GFP-tagged Rab11 (wt or S25N) and FLAG-tagged Msps were cotransfected into S2 cells and immunoprecipitated with FLAG beads. Inputs and immunoprecipitates (IP) were blotted with the indicated antibodies. Sizes of molecular weight markers in kiloDalton (kD) are shown. IgG denotes antibody heavy chains. H Patronin localizes along dendrites and in dendritic tips in first instar c4da neurons. The side panel shows an enlarged image of a dendrite (boxed area in larger image). The asterisk denotes the position of the soma. Scale bars are 5 μm in A and D, 2 μm in B and 10 μm in G.

The microtubule polymerase Msps/XMAP215 is another potential MTOC component linked to pruning and dendritic microtubule organization (Tang et al, 2020). In co- immunoprecipitations from transfected S2 cells, GFP-tagged Rab11, but not GTP- binding-deficient Rab11 S25N, could be co-precipitated with FLAG-tagged Msps (Fig. 6 G). Lastly, the minus end-binder Patronin is required for dendrite pruning and localizes to growing dendrite tips in PVD neurons (Wang et al, 2019; He et al, 2020). Transgenic Patronin::GFP was also enriched at dendrite tips - as well as in streaks along dendrite shafts - in first instar c4da neurons (Fig. 6 H). Taken together, our data suggest the existence of a developmentally regulated, Rab11-based MTOC in early growing c4da neuron dendrites that likely also acts at tips.

Shot recruits EB1 to tips and interacts with candidate MTOC components

The above observations opened up the possibility that Shot is part of an early developmental dendritic tip MTOC. In support of this notion, loss of Shot already caused altered microtubule orientation at the first instar stage (Fig. 7 A). To test whether Shot can recruit the microtubule nucleation machinery, we coexpressed Shot::GFP with red fluorescently tagged EB1 in c4da neurons. In the presence of Shot::GFP, EB1 also became recruited to dendrite tips at the first instar and colocalized with Shot::GFP (Fig. 7 B, C), furthermore, EB1 comets could be seen emanating from these Shot/EB1-rich tips (Fig. 7 D). To link Shot to the other microtubule regulators implicated in early dendritic microtubule nucleation, we tested for genetic interactions using pruning as a readout. To this end, we knocked down shot in c4da neurons and then crossed in heterozygous P element mutants in Patronin and Msps, or a dsRNA construct targeting EB1. While these manipulations on their own did not cause pruning defects, the number of secondary and higher order dendrite branches attached to the soma upon was strongly increased when combined with shot knockdown, indicating phenotypic enhancement (Fig. 7 E - I). To assess genetic interactions with Rab11, we knocked down rab11, which by itself already causes highly penetrant pruning defects, but most higher order dendrites are still pruned and only primary dendrites remain (Krämer et al, 2019) (Fig. 7 J, L). In contrast, rab11 knockdown in a shot3/+ heterozygous background also caused significantly more higher order dendrites to stay attached to the soma than with than rab11 knockdown alone (Fig. 7 K, L). Thus, EB1 imaging and genetic interactions link Shot to dendritic MTOC functions.

Evidence that Shot acts as part of a dendritic MTOC.

A Frequency of anterograde EB1 comets at the L1 stage in control (Orco dsRNA) and shot knockdown neurons (shot dsRNA, shot3/+). * P<0.05, Mann Whitney U test. B Shot recruits EB1 to dendrite tips. EB1::mScarlet3 was expressed in c4da neurons without (upper panel) or with ShotFL::GFP (lower panel) and visualized at the first instar larval stage. The position of the dendrite tip is indicated by an asterisk. C Fluorescence intensity profiles of EB1::mScarlet3 in the distal 10 μm of first instar dendrites. Solid lines indicate average, envelopes indicate S. D. (N=9 each). C’ Table showing significance between genotypes for the indicated distances from the tip. **** P<0.0001, ** P<0.01, Mann Whitney U test. D Example kymograph of EB1::mScarlet3 and Shot::GFP at a dendritic tip. Tip position is to the left, and arrows in the EB1::mScarlet3 and merge panels indicate a comet originating there. E - L Synergistic genetic interactions between shot and other dendritic microtubule orientation/MTOC factors during c4da neuron dendrite pruning. E - I shot dsRNA was expressed in c4da neurons of animals in the indicated backgrounds, and pruning defects were quantified at 16 h APF. E C4da neuron expressing shot dsRNA. F C4da neuron expressing shot dsRNA in a msps/+ background. G C4da neuron expressing shot dsRNA in a patronin/+ background. H C4da neuron co-expressing shot and EB1 dsRNAs. I Severity of pruning defects in E - G. N = 33 - 52, ** P<0.01, *** P<0.001, Wilcoxon’s test. J, K rab11 dsRNA was expressed in c4da neurons of control animals (J) or in a shot3/+ heterozygous background (K), and pruning defects were quantified at 16 h APF. L Severity of pruning defects in J, K. N = 41 and 42. **** P<0.0001, Wilcoxon’s test. Scale bars are 5 μm in B, 2 μm in D and 50 μm in E.

Discussion

Microtubule organization in neurites is crucial for neuronal polarity, but the developmental mechanisms are still emerging. Here, we show that the spectraplakin Shot is required for the uniform plus end-in organization of dendritic microtubules. Our data indicate that the role of Shot is twofold. On the one hand, we provide evidence that Shot anchors microtubules via actin in mature neurons. If unanchored, a subset of plus end-out oriented microtubules in Shot loss-of-function neurons can undergo sliding, which has been linked to microtubule misorientation (He et al, 2020). The observation that only plus end-out microtubules show this behavior suggests that there might be other, orientation-dependent anchoring mechanisms that do not depend on Shot. Surprisingly, overexpression of the GAS2 microtubule binding domain was sufficient to rescue the phenotypes of shot mutant neurons. Microtubule length regulation and catastrophe rescue have recently been shown to be important for maintenance of dendritic microtubule orientation (Liang et al, 2024). It will be interesting to test whether the GAS2 domain can fulfil a similar function in c4da neurons.

The other function of Shot is as part of an early dendritic MTOC. Similar to the C. elegans PVD neuron, c4da neurons have a transient microtubule-linked structure at the tip of early growing dendrites, and they depend on Rab11 for proper microtubule organization. Shot interacts genetically with Rab11 and is a component of this tip structure. Early-stage tip MTOCs and Rab11 dependence may therefore be a general feature of dendritic development. The exact function of Shot in the tip MTOC is still unclear, but the observation that Shot can recruit EB1 to tips suggests that it may recruit several MTOC components. The dendritic tip MTOCs of C. elegans were suggested to be transported to dendrite tips on transient plus end-out microtubules (Liang et al, 2020). Plus end-out microtubules can be observed in terminal c4da neuron dendrites (Ori-McKenney et al, 2012; Hu et al, 2022). It is interesting to speculate that Shot might participate in such a transport function. Crucially, both proposed functions of Shot during microtubule organization depend on actin binding, linking the two main cytoskeletal components in neurite organization.

Materials and methods

Fly Strains

C4da neurons were labeled by UAS-CD8::GFP or UAS-tdtomato expression under ppk-GAL4 (Grueber et al, 2007). MARCM clones of shot3mutants (BL 5141) were induced with SOP-FLP (Matsubara et al, 2011) and labeled by tdtomato expression under nsyb-GAL4R57C10. Endogenously tagged Shot::GFP was from (Voelzmann et al, 2024). EB1 analyses were carried out using either UAS-EB1::GFP (Zheng et al, 2008), ppk-EB1::GFP (Arthur et al, 2015), or UAS-EB1::mScarlet3. The following dsRNA lines were used: Orco (BL 31278) as control, Shot (BL 28336), Rab11 (VDRC 22198), Rab5 (VDRC 34096), EB1 (VDRC 24451). All dsRNAs were coexpressed with UAS-dcr2 (Dietzl et al, 2007). UAS transgenes were UAS-ShotRE::GFP on the X chromosome (BL 29044), or mobilized to the third chromosome (this study), UAS-ShotRC::GFP (1CH1, BL 29042), UAS-Shot Cterm::YFP (Nashchekin et al, 2024), UAS-kin::lacZ (Clark et al, 1997), UAS-Mical (Terman et al, 2002), UAS-αtub84B::mCherry (Villars et al, 2022), UAS-Rab11::GFP (BL 8506), UAS-tauHA (Herzmann et al, 2017), UAS-γtub23C::GFP (Nguyen et al, 2011), UAS-Patronin::GFP (Derivery et al, 2015), UAS- cytoABKAR TA (Marzano et al, 2021) as inert UAS control. Additional mutant alleles were mspsMI14162 (BL 59478), patronink07433 (Kyoto 111217).

Cloning and transgenes

Shot sgRNAs GCTGCCCTCTCAGGCCGATT (target 1) and GAGTTCTCCAGAGTGGTCAC (target 2) were cloned into pCFD4w+ (gift from S. Schirmeier). For transgenesis, plasmids were injected into flies carrying the 86Fb acceptor site. C-terminally GFP-tagged Rab11 for biochemistry was cloned into pUAST attB carrying a C-terminal GFP tag using TOPO/TA cloning. The S25N mutation was inserted by PCR mutatgenesis. For Rab11::mCherry, the Rab11 entry clone was used to generate pUAST attB Rab11::mCherry, the insert was subcloned into pUAST, which was then injected into w1118 flies using classical transposase- mediated transgenesis. To generate EB1::mScarlet3, a Drosophila codon optimized mScarlet3 (Gaudella et al, 2023, synthesized by Twist Bioscience) was added after the EB1 coding sequence separated by a CACACCTCCACTACCGCTAGCAGGCCGGCCACGCGTGGTACCTTCTGGTCCA linker and cloned into pUAST. The resulting plasmid was injected into wild type embryos by Bestgene. C-terminally FLAG-tagged Msps (isoform RB) was cloned into pUAST attB using standard cloning procedures.

Immunofluorescence

Briefly, appropriately staged pupal filets were fixed in 4% formaldehyde, blocked in PBS with 0.3 % Triton X-100 and 10% goat serum, and incubated with antibody in blocking buffer over night. C4da neurons labeled by ppk promotor fusions, endogenous or transgenic GFP-tagged Shot were visualized with chicken (1:500, Aves labs), rabbit (1:1000, Invitrogen A11122) or mouse (1:1000, Invitrogen A1120) anti-GFP. Tdtomato and mcherry-tagged transgenes were detected with rat anti-mCherry (1:1000, Invitrogen) or rabbit anti-DsRed antibodies (1:1000, Clontech). Kin::β-galactosidase fusion proteins were detected with rabbit anti-β-galactosidase (Cappel, preabsorbed in-house, 1:700). Secondary antibodies for regular immunofluorescence were conjugated to Alexa Fluor 488, 568 or 647. For STED microscopy of tagged tubulin, Abberior anti rabbit STAR RED were used as secondary antibodies.

Microscopy, time lapse imaging

For analyses of dendrite pruning phenotypes at 16 h APF, animals were dissected out of the pupal case and dorsal ddaC c4da neurons in segments A2 - A5 were imaged live on a Zeiss LSM710 confocal microscope with a 20x Plan Apochromat water objective (1.0 NA). For regrowth analyses at 72 h APF, animals were dissected out of the pupal case and ventral v’ada c4da neurons in segments A2 - A4 were imaged as described (Sanal et al, 2023). For live protein localization analyses, first instar larvae were immobilized on double sided tape, third instar larvae were briefly anaesthesized with ether before imaging. Neurons were then imaged live on Zeiss LSM710 or LSM880 microscopes with 63x Plan Apochromat DIC M27 oil objectives (1.4 NA). EB1::GFP imaging was performed on a Zeiss LSM880 microscope using a 40x Plan Apochromat FCS M27 (1.2 NA) oil objective. Consecutive images of a single plane were taken every second for 1 - 2 minutes.

Immunofluorescence images of kin::lacZ were obtained on a Zeiss LSM 710 microscope using a 40x Zeiss C Apochromat water objective (1.1 NA). Structured illumination microscopy (SIM) was performed on a Zeiss Elyra 7 microscope with a 63x Plan Apochromat DIC M27 oil objective (1.4 NA) with an Edge 4.2 SIM camera (50 ms exposure time), and deconvolution was done using SIM2 software. All microscopes used ZEN Black software. Stimulated Emission Depletion Microscopy (STED) images were taken with an abberior STEDYCON equipped with an inverted IX83 microscope (Olympus), a 100x oil objective (Olympus UPLXAPO100XO, NA 1.45), using pulsed excitation lasers at 640 nm and a pulsed STED laser operating at 775 nm, continuous autofocus and gated detection with avalanche photodiode element detectors. All acquisition operations were controlled by STEDYCON Software (abberior Instruments), and images were deconvolved using Huygens deconvolution software. Microscopic images shown are maximum projections or (where indicated) single plane images. All processing was done in Fiji (Schindelin et al, 2012) using the plug-ins NeuronJ for dendrite length measurements, and Image Stabilizer for EB1::GFP comet analysis. Kymographs from different channels were overlayed using the KymoResliceWide plugin.

S2 cell culture, immunoprecipitation and Western blots

pUAST Rab11::GFP (wt or S25N) was cotransfected into S2 cells with pUAST- MspsFLAG and Actin5C-GAL4 as described (Herzmann et al, 2017). After 72 hours, cells were harvested in ice-cold PBS, and lysed in cold lysis buffer (100 mM NaCl, 5 mM MgCl2, 50 mM Tris/HCl pH 7.6, 5 % glycerol, 1 % Triton X-100, 1x complete protease inhibitor). Cleared lysates were incubated with anti-FLAG beads (Sigma A- 2220) for 2 hours. After three washes with lysis buffer, bound proteins were eluted in SDS sample buffer. Samples were run on 8 % gels and blotted with JL-8 anti-GFP (Clontech, 1:1000) and FLAG M2 (Sigma F3165, 1:5000) antibodies. Bands were visualized on an Amersham Imager 680.

Quantification and statistical analyses

Phenotypic penetrance was assessed by counting the number of neurons with dendrites still attached to the soma. Here, significance was determined using a categorical two-tailed Fisher’s exact test (graphpad.com). Length of unpruned dendrites were measured using the Fiji NeuronJ plugin and compared using the Wilcoxon Mann Whitney test (Marx et al., 2016).

In order to quantify Shot dendrite tip localization, intensity profiles of the distal 10 μm of each dendrite were averaged. The resulting curves were then sectioned into 2 μm segments for statistical comparisons between genotypes using the Mann Whitney test.

eLife Assessment

Davies et al. present a valuable study proposing that Shot can act as a molecular linker between microtubules and actin during dendrite pruning, suggesting an intriguing role in non-centrosomal microtubule organization. However, the experimental evidence is incomplete and does not robustly support these claims, and the lack of a cohesive model connecting the findings weakens the overall impact. While the data suggest that Shot, actin, and microtubule nucleation contribute to dendritic pruning, their precise interplay remains unresolved.

Reviewer #1 (Public review):

Summary:

The Neuronal microtubule cytoskeleton is essential long long-range transport in axons and dendrites. The axon-specific plus-end out microtubule organization vs the dendritic-specific plus-end in organization allows for selective transport into each neurite, setting up neuronal polarity. In addition, the dendritic microtubule organization is thought to be important for dendritic pruning in Drosophila during metamorphosis. However, the precise mechanisms that organize microtubules in neurons are still incompletely understood.

In the current manuscript, the authors describe the spectraplakin protein Shot as important in developmental dendritic pruning. They find that Shot has dendritic microtubule polarity defects, which, based on their rescues and previous work, is likely the reason for the pruning defect.

Since Shot is a known actin-microtubule crosslinker, they also investigate the putative role of actin and find that actin is also important for dendritic pruning. Finally, they find that several factors that have been shown to function as a dendritic MTOC in C. elegans also show a defect in Drosophila upon depletion.

Strengths:

Overall, this work was technically well-performed, using advanced genetics and imaging. The author reports some interesting findings identifying new players for dendritic microtubule organization and pruning.

Weaknesses:

The evidence for Shot interacting with actin for its functioning is contradictory. The Shot lacking the actin interaction domain did not rescue the mutant; however, it also has a strong toxic effect upon overexpression in wildtype (Figure S3), so a potential rescue may be masked. Moreover, the C-terminus-only construct, which carries the GAS2-like domain, was sufficient to rescue the pruning. This actually suggests that MT bundling/stabilization is the main function of Shot (and no actin binding is needed). On the other hand, actin depolymerization leads to some microtubule defects and subtle changes in shot localization in young neurons (not old ones). More importantly, it did not enhance the microtubule or pruning defects of the Shot domain, suggesting these act in the same pathway. Interesting to note is that Mical expression led to microtubule defects but not to pruning defects. This argues that MT organization effects alone are not enough to cause pruning defects. This may be be good to discuss. For the actin depolymerization, the authors used overexpression of the actin-oxidizing Mical protein. However, Mical may have another target. It would be good to validate key findings with better characterized actin targeting tools.

In analogy to C. elegans, where RAB-11 functions as a ncMTOC to set up microtubules in dendrites, the authors investigated the role of these in Drosophila. Interestingly, they find that rab-11 also colocalizes to gamma tubulin and its depletion leads to some microtubule defects. Furthermore, they find a genetic interaction between these components and Shot; however, this does not prove that these components act together (if at all, it would be the opposite). This should be made more clear. What would be needed to connect these is to address RAB-11 localization + gamma-tubulin upon shot depletion.

All components studied in this manuscript lead to a partial reversal of microtubules in the dendrite. However, it is not clear from how the data is represented if the microtubule defect is subtle in all animals or whether it is partially penetrant stronger effect (a few animals/neurons have a strong phenotype). This is relevant as this may suggest that other mechanisms are also required for this organization, and it would make it markedly different from C. elegans. This should be discussed and potentially represented differently.

Reviewer #2 (Public review):

Summary:

In their manuscript, the authors reveal that the spectraplakin Shot, which can bind both microtubules and actin, is essential for the proper pruning of dendrites in a developing Drosophila model. A molecular basis for the coordination of these two cytoskeletons during neuronal development has been elusive, and the authors' data point to the role of Shot in regulating microtubule polarity and growth through one of its actin-binding domains. The authors also propose an intriguing new activity for a spectraplakin: functioning as part of a microtubule-organizing center (MTOC).

Strengths:

(1) A strength of the manuscript is the authors' data supporting the idea that Shot regulates dendrite pruning via its actin-binding CH1 domain and that this domain is also implicated in Shot's ability to regulate microtubule polarity and growth (although see comments below); these data are consistent with the authors' model that Shot acts through both the actin and microtubule cytoskeletons to regulate neuronal development.

(2) Another strength of the manuscript is the data in support of Rab11 functioning as an MTOC in young larvae but not older larvae; this is an important finding that may resolve some debates in the literature. The finding that Rab11 and Msps coimmunoprecipitate is nice evidence in support of the idea that Rab11(+) endosomes serve as MTOCs.

Weaknesses:

(1) A significant, major concern is that most of the authors' main conclusions are not (well) supported, in particular, the model that Shot functions as part of an MTOC. The story has many interesting components, but lacks the experimental depth to support the authors' claims.

(2) One of the authors' central claims is that Shot functions as part of a non-centrosomal MTOC, presumably a MTOC anchored on Rab11(+) endosomes. For example, in the Introduction, last paragraph, the authors summarize their model: "Shot localizes to dendrite tips in an actin-dependent manner where it recruits factors cooperating with an early-acting, Rab11-dependent MTOC." This statement is not supported. The authors do not show any data that Shot localizes with Rab11 or that Rab11 localization or its MTOC activity is affected by the loss of Shot (or otherwise manipulating Shot). A genetic interaction between Shot and Rab11 is not sufficient to support this claim, which relies on the proteins functioning together at a certain place and time. On a related note, the claim that Shot localization to dendrite tips is actin-dependent is not well supported: the authors show that the CH1 domain is needed to enrich Shot at dendrite tips, but they do not directly manipulate actin (it would be helpful if the authors showed the overexpression of Mical disrupted actin, as they predict).

(3) The authors show an image that Shot colocalizes with the EB1-mScarlet3 comet initiation sites and use this representative image to generate a model that Shot functions as part of an MTOC. However, this conclusion needs additional support: the authors should quantify the frequency of EB1 comets that originate from Shot-GFP aggregates, report the orientation of EB1 comets that originate from Shot-GFP aggregates (e.g., do the Shot-GFP aggregates correlate with anterogradely or retrogradely moving EB1 comets), and characterize the developmental timing of these events. The genetic interaction tests revealing ability of shot dsRNA to enhance the loss of microtubule-interacting proteins (Msps, Patronin, EB1) and Rab11 are consistent with the idea that Shot regulates microtubules, but it does not provide any spatial information on where Shot is interacting with these proteins, which is critical to the model that Shot is acting as part of a dendritic MTOC.

(4) It is unclear whether the authors are proposing that dendrite pruning defects are due to an early function of Shot in regulating microtubule polarity in young neurons (during 1st instar larval stages) or whether Shot is acting in another way to affect dendrite pruning. It would be helpful for the authors to present and discuss a specific model regarding Shot's regulation of dendrite pruning in the Discussion.

(5) The authors argue that a change in microtubule polarity contributes to dendrite pruning defects. For example, in the Introduction, last paragraph, the authors state: "Loss of Shot causes pruning defects caused by mixed orientation of dendritic microtubules." The authors show a correlative relationship, not a causal one. In Figure 4, C and E, the authors show that overexpression of Mical disrupts microtubule polarity but not dendrite pruning, raising the question of whether disrupting microtubule polarity is sufficient to cause dendrite pruning defects. The lack of an association between a disruption in microtubule polarity and dendrite pruning in neurons overexpressing Mical is an important finding.

(6) The authors show that a truncated Shot construct with the microtubule-binding domain, but no actin-binding domain (Shot-C-term), can rescue dendrite pruning defects and Khc-lacZ localization, whereas the longer Shot construct that lacks just one actin-binding domain ("delta-CH1") cannot. Have the authors confirmed that both proteins are expressed at equivalent levels? Based on these results and their finding that over-expression of Shot-delta-CH1 disrupts dendrite pruning, it seems possible that Shot-delta-CH1 may function as a dominant-negative rather than a loss-of-function. Regardless, the authors should develop a model that takes into account their findings that Shot, without any actin-binding domains and only a microtubule-binding domain, shows robust rescue.

(7) The authors state that: "The fact that Shot variants lacking the CH1 domain cannot rescue the pruning defects of shot[3] mutants suggested that dendrite tip localization of Shot was important for its function." (pages 10-11). This statement is not accurate: the Shot C-term construct, which lacks the CH1 domain (as well as other domains), is able to rescue dendrite pruning defects.

(8) The authors state that: "In further support of non-functionality, overexpression of Shot[deltaCH1] caused strong pruning defects (Fig. S3)." (page 8). Presumably, these results indicate that Shot-delta-CH1 is functioning as a dominant-negative since a loss-of-function protein would have no effect. The authors should revise how they interpret these results. This comment is related to another comment about the ability of Shot constructs to rescue the shot[3] mutant.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

The Neuronal microtubule cytoskeleton is essential for long-range transport in axons and dendrites. The axon-specific plus-end out microtubule organization vs the dendritic-specific plus-end in organization allows for selective transport into each neurite, setting up neuronal polarity. In addition, the dendritic microtubule organization is thought to be important for dendritic pruning in Drosophila during metamorphosis. However, the precise mechanisms that organize microtubules in neurons are still incompletely understood.

In the current manuscript, the authors describe the spectraplakin protein Shot as important in developmental dendritic pruning. They find that Shot has dendritic microtubule polarity defects, which, based on their rescues and previous work, is likely the reason for the pruning defect.

Since Shot is a known actin-microtubule crosslinker, they also investigate the putative role of actin and find that actin is also important for dendritic pruning. Finally, they find that several factors that have been shown to function as a dendritic MTOC in C. elegans also show a defect in Drosophila upon depletion.

Strengths:

Overall, this work was technically well-performed, using advanced genetics and imaging. The author reports some interesting findings identifying new players for dendritic microtubule organization and pruning.

We thank reviewer 1 for this assessment.

Weaknesses:

(1) The evidence for Shot interacting with actin for its functioning is contradictory. The Shot lacking the actin interaction domain did not rescue the mutant; however, it also has a strong toxic effect upon overexpression in wildtype (Figure S3), so a potential rescue may be masked. Moreover, the C-terminus-only construct, which carries the GAS2-like domain, was sufficient to rescue the pruning. This actually suggests that MT bundling/stabilization is the main function of Shot (and no actin binding is needed).

Thank you for this comment. We agree that the rescue and overexpression experiments with UAS-Shot ΔCH1 (lacking only the first actin-binding domain) and UAS-Shot CTerm (containing only the MT binding domain) allow for other interpretations than ours that both features are necessary for Shot function in dendrites. To address this issue, we tested two more Shot domain deletion mutants, Shot ΔABD (lacking both actin-binding domains) and Shot ΔCTail, which lacks the MT binding domain (Figure 3). Neither of the latter two constructs can rescue the pruning and MT orientation phenotypes of shot mutants, confirming that actin binding is required for these functions in the context of full-length Shot. Importantly, in contrast to the previously used Shot ΔCH1, neither Shot ΔABD nor Shot ΔCTail inhibit pruning upon overexpression, ruling out other interpretations of overexpression toxicity (Fig. S2). The observation that Shot CTerm (containing only the MT binding domain) is sufficient to rescue indicates that actin binding is only required for Shot function in the context of full-length Shot. Interestingly, this is reminiscent of a proposed autoregulation mechanism where actin binding disinhibits Shot's MT binding domain (Applewhite 2013).

Figure 3, Figure S2 (Shot domain analysis): We added two new constructs (Shot ΔABD, Shot ΔCtail) and a Shot BAC as additional control in Figure 3.

(2) On the other hand, actin depolymerization leads to some microtubule defects and subtle changes in shot localization in young neurons (not old ones). More importantly, it did not enhance the microtubule or pruning defects of the Shot domain, suggesting these act in the same pathway. Interesting to note is that Mical expression led to microtubule defects but not to pruning defects. This argues that MT organization effects alone are not enough to cause pruning defects. This may be good to discuss.

Thank you for this comment. We think that microtubule orientation defects are tightly linked to dendrite pruning defects, as all known conditions causing orientation defects either cause pruning defects, or they act as genetic enhancers of pruning mutants. We now clarify this in the introduction citing the example of EB1 RNAi (causes orientation defects (Matties et al., 2010), enhances pruning mutants (Herzmann et al., 2018)). This is also the case for Mical overexpression, which causes orientation defects and enhances the pruning defects caused by Shot RNAi. To make this point clearer, we now also show representative images of pruning defects caused by either Shot RNAi alone or Shot RNAi combined with Mical overexpression.

Introduction - we explain better the genetic relationship between microtubule orientation and pruning. Figure 4 (phenotypic enhancement of shot knockdown by Mical overexpression): added representative images.

(3) For the actin depolymerization, the authors used overexpression of the actin-oxidizing Mical protein. However, Mical may have another target. It would be good to validate key findings with better characterized actin targeting tools.

Thank you for this comment. To our knowledge, there are no good tools to globally manipulate actin dynamics in vivo in a similar manner to, e. g., latrunculin in cultured cells. According to the literature, Micals are highly specific for actin. To enhance our analysis, we assessed the effect of Mical overexpression on actin distribution in c4da neurons (as assessed by LifeAct-GFP) and found significant effects at the first instar.

Added new analysis of effect of Mical overexpression on Lifeact::GFP distribution in c4da neurons (new Fig. S3), added references for Mical targets.

(4) In analogy to C. elegans, where RAB-11 functions as a ncMTOC to set up microtubules in dendrites, the authors investigated the role of these in Drosophila. Interestingly, they find that rab-11 also colocalizes to gamma tubulin and its depletion leads to some microtubule defects. Furthermore, they find a genetic interaction between these components and Shot; however, this does not prove that these components act together (if at all, it would be the opposite). This should be made more clear. What would be needed to connect these is to address RAB-11 localization + gamma-tubulin upon shot depletion.

All components studied in this manuscript lead to a partial reversal of microtubules in the dendrite. However, it is not clear from how the data is represented if the microtubule defect is subtle in all animals or whether it is partially penetrant stronger effect (a few animals/neurons have a strong phenotype). This is relevant as this may suggest that other mechanisms are also required for this organization, and it would make it markedly different from C. elegans. This should be discussed and potentially represented differently.

Thank you for this comment. We agree that the genetic interaction with pruning as readout does not prove that Rab11 and Shot act in the exact same pathway during establishment of microtubule organization. To address this question, we live-imaged Rab11::GFP vesicles in control and Shot knockdown neurons and found no significant difference in motile behavior (new Figure 7). As Rab11::GFP also is not visibly enriched at dendrite tips at the first instar stage, we cannot confidently state that a Rab11 tip MTOC exists in Drosophila neurons. Still, we only see EB1 comets originating from Rab11 puncta at the first instar, but not the third. We therefore rephrased and toned down our interpretation and state now that Rab11 may be a component of a developmentally transient MTOC but likely acts independently of Shot.

Unfortunately, we do not see many comets per neuron with our fluorescently tagged EB1 constructs, indicating that only a fraction of them are labeled. This makes it very difficult to judge severity between single neurons. The impression is that usually only some comets per neuron have reversed orientation, which would be consistent with several independent mechanisms for orientation.

Reviewer #2 (Public review):

Summary:

In their manuscript, the authors reveal that the spectraplakin Shot, which can bind both microtubules and actin, is essential for the proper pruning of dendrites in a developing Drosophila model. A molecular basis for the coordination of these two cytoskeletons during neuronal development has been elusive, and the authors' data point to the role of Shot in regulating microtubule polarity and growth through one of its actin-binding domains. The authors also propose an intriguing new activity for a spectraplakin: functioning as part of a microtubule-organizing center (MTOC).

Strengths:

(1) A strength of the manuscript is the authors' data supporting the idea that Shot regulates dendrite pruning via its actin-binding CH1 domain and that this domain is also implicated in Shot's ability to regulate microtubule polarity and growth (although see comments below); these data are consistent with the authors' model that Shot acts through both the actin and microtubule cytoskeletons to regulate neuronal development.

(2) Another strength of the manuscript is the data in support of Rab11 functioning as an MTOC in young larvae but not older larvae; this is an important finding that may resolve some debates in the literature. The finding that Rab11 and Msps coimmunoprecipitate is nice evidence in support of the idea that Rab11(+) endosomes serve as MTOCs.

Thank you for these comments.

Weaknesses:

(1) A significant, major concern is that most of the authors' main conclusions are not (well) supported, in particular, the model that Shot functions as part of an MTOC. The story has many interesting components, but lacks the experimental depth to support the authors' claims.

Thank you for this comment. We agree that we cannot conclusively state that Shot is part of a classical MTOC where microtubules are nucleated and therefore tone down our conclusions regarding this.

(2) One of the authors' central claims is that Shot functions as part of a non-centrosomal MTOC, presumably a MTOC anchored on Rab11(+) endosomes. For example, in the Introduction, last paragraph, the authors summarize their model: "Shot localizes to dendrite tips in an actin-dependent manner where it recruits factors cooperating with an early-acting, Rab11-dependent MTOC." This statement is not supported. The authors do not show any data that Shot localizes with Rab11 or that Rab11 localization or its MTOC activity is affected by the loss of Shot (or otherwise manipulating Shot). A genetic interaction between Shot and Rab11 is not sufficient to support this claim, which relies on the proteins functioning together at a certain place and time. On a related note, the claim that Shot localization to dendrite tips is actin-dependent is not well supported: the authors show that the CH1 domain is needed to enrich Shot at dendrite tips, but they do not directly manipulate actin (it would be helpful if the authors showed the overexpression of Mical disrupted actin, as they predict).

Thank you for these comments. In response, we tested whether Shot knockdown affects Rab11 vesicle motility in first instar dendrites and find that this is likely not the case (new Fig. 7 K, L). We therefore remove the proposal that Shot directly regulates Rab11. Regarding the actin dependence of Shot tip localization, we based this conclusion on the observations that Shot ΔCH1 does not localize to tips, and that Mical overexpression significantly broadens the Shot tip signal (Fig. 5 E, F, J - previously Fig. 5 C, D, G). To further deepen this conclusion, we now add a localization analysis of a Shot mutant lacking both actin-binding CH domains (Shot ΔABD) and find that this also abrogates tip enrichment.

(3) The authors show an image that Shot colocalizes with the EB1-mScarlet3 comet initiation sites and use this representative image to generate a model that Shot functions as part of an MTOC. However, this conclusion needs additional support: the authors should quantify the frequency of EB1 comets that originate from Shot-GFP aggregates, report the orientation of EB1 comets that originate from Shot-GFP aggregates (e.g., do the Shot-GFP aggregates correlate with anterogradely or retrogradely moving EB1 comets), and characterize the developmental timing of these events. The genetic interaction tests revealing ability of shot dsRNA to enhance the loss of microtubule-interacting proteins (Msps, Patronin, EB1) and Rab11 are consistent with the idea that Shot regulates microtubules, but it does not provide any spatial information on where Shot is interacting with these proteins, which is critical to the model that Shot is acting as part of a dendritic MTOC.

Thank you for these comments. Due in part to low overall EB1 comet density in c4da neurons, we could rarely see comets originating from tip-localized Shot::GFP at the first instar stage - one example is shown in new Fig. S5C (previously Fig. 7D). At later stages, Shot is relatively evenly localized along the dendritic plasma membrane. As we cannot be certain about the relatively vague MTOC conclusions, we decided to rephrase our conclusions and state - based on genetic and functional data - that Shot's ability to stabilize microtubules is required, likely at dendritic tips.

(4) It is unclear whether the authors are proposing that dendrite pruning defects are due to an early function of Shot in regulating microtubule polarity in young neurons (during 1st instar larval stages) or whether Shot is acting in another way to affect dendrite pruning. It would be helpful for the authors to present and discuss a specific model regarding Shot's regulation of dendrite pruning in the Discussion.

Thank you for these comments. It is indeed our hypothesis that Shot's early role in setting up microtubule organization is also crucial for pruning, this is the most parsimonious explanation as pruning and orientation defects always go hand in hand. We added the following sentence in the Discussion: "As we do not have evidence for Shot regulation at the onset of the pupal stage (Fig. S4), the function of Shot crucial for pruning is most arguably its early role in setting up uniform microtubule orientation."

(5) The authors argue that a change in microtubule polarity contributes to dendrite pruning defects. For example, in the Introduction, last paragraph, the authors state: "Loss of Shot causes pruning defects caused by mixed orientation of dendritic microtubules." The authors show a correlative relationship, not a causal one. In Figure 4, C and E, the authors show that overexpression of Mical disrupts microtubule polarity but not dendrite pruning, raising the question of whether disrupting microtubule polarity is sufficient to cause dendrite pruning defects. The lack of an association between a disruption in microtubule polarity and dendrite pruning in neurons overexpressing Mical is an important finding.

Thank you for this comment. Microtubule orientation defects are tightly linked to dendrite pruning defects, as all known conditions causing orientation defects either cause pruning defects, or they act as genetic enhancers of pruning mutants. We now clarify this in the introduction citing the example of EB1 RNAi (which causes orientation defects (Matties et al., 2010), and does not cause pruning defects by itself, but enhances the effects of other pruning mutants (Herzmann et al., 2018)). This is also the case for Mical overexpression, which causes orientation defects and enhances the pruning defects caused by Shot RNAi. To make this point clearer, we now also show representative images of pruning defects caused by either Shot RNAi alone or Shot RNAi combined with Mical overexpression.

Introduction - we explain better the genetic relationship between microtubule orientation and pruning. Figure 4 (phenotypic enhancement of shot knockdown by Mical overexpression): added representative images

(6) The authors show that a truncated Shot construct with the microtubule-binding domain, but no actin-binding domain (Shot-C-term), can rescue dendrite pruning defects and Khc-lacZ localization, whereas the longer Shot construct that lacks just one actin-binding domain ("delta-CH1") cannot. Have the authors confirmed that both proteins are expressed at equivalent levels? Based on these results and their finding that over-expression of Shot-delta-CH1 disrupts dendrite pruning, it seems possible that Shot-delta-CH1 may function as a dominant-negative rather than a loss-of-function. Regardless, the authors should develop a model that takes into account their findings that Shot, without any actin-binding domains and only a microtubule-binding domain, shows robust rescue.

Thank you for this constructive criticism. We agree that the rescue and overexpression experiments with UAS-Shot ΔCH1 (lacking only the first actin-binding domain) and UAS-Shot CTerm (containing only the MT binding domain) allow for other interpretations than ours that both features are necessary for Shot function in dendrites. To address this issue, we tested two more Shot domain deletion mutants, Shot ΔABD, lacking both actin-binding domains, and Shot ΔCTail, which lacks the MT binding domain (Figure 3). Neither of these two constructs can rescue the pruning and MT orientation phenotypes of shot mutants, confirming that both actin and microtubule binding are required for these functions in the context of full-length Shot. Importantly, in contrast to the previously tested Shot ΔCH1, neither Shot ΔABD nor Shot ΔCTail inhibit pruning upon overexpression, ruling out other interpretations of overexpression toxicity. The observation that Shot CTerm is sufficient to rescue indicates that actin binding is only required for Shot function in the context of full-length Shot. Interestingly, this is reminiscent of a proposed autoregulation mechanism where actin binding disinhibits Shot's MT binding domain (Applewhite 2013).

Figure 3, Figure S2 (Shot domain analysis): We added two new constructs (Shot ΔABD, Shot ΔCTail) and a Shot BAC as additional control in Figure 3.

(7) The authors state that: "The fact that Shot variants lacking the CH1 domain cannot rescue the pruning defects of shot[3] mutants suggested that dendrite tip localization of Shot was important for its function." (pages 10-11). This statement is not accurate: the Shot C-term construct, which lacks the CH1 domain (as well as other domains), is able to rescue dendrite pruning defects.

Thank you for this comment. Shot ΔCH1 indeed does show a partial rescue of the pruning defects (but not of the polarity defects). To clarify whether actin binding is required for Shot function in dendrites, we tested the additional Shot construct Shot ΔABD and found that this construct could not rescue the defects at all (Fig. 3).

Fig. 3, added Shot ΔABD.

(8) The authors state that: "In further support of non-functionality, overexpression of Shot[deltaCH1] caused strong pruning defects (Fig. S3)." (page 8). Presumably, these results indicate that Shot-delta-CH1 is functioning as a dominant-negative since a loss-of-function protein would have no effect. The authors should revise how they interpret these results. This comment is related to another comment about the ability of Shot constructs to rescue the shot [3] mutant.

Thank you for this comment. We agree that the rescue and overexpression experiments with UAS-Shot ΔCH1 (lacking only the first actin-binding domain) allow for other interpretations than ours - that actin binding is necessary for Shot function in dendrites. To address this issue, we tested another Shot domain deletion mutant, Shot ΔABD (lacking both actin-binding domains) (Figure 3). This construct did not rescue the pruning and MT orientation phenotypes of shot mutants, and also does not cause pruning defects upon overexpression, confirming that actin binding is required in the context of full-length Shot. As our domain analysis is reminiscent of a proposed autoregulation mechanism where actin binding disinhibits Shot's MT binding domain (Applewhite 2013), it is interesting to speculate that the overexpression toxicity of Shot ΔCH1 may stem from a residual ability of the second CH domain to relieve autoinhibition.

Fig. 3, Fig. S2, added Shot ΔABD and Shot ΔCTail to domain analyses.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) Not enough info was given about how the statistics were performed in e.g., Figures 5E, 5H and 7C. Many data are highly significant, but this is not always directly clear to me from the images, e.g., 7C 6-10um. Or comparing 7I to 7L, how come the significance is so much different

Thank you for this comment. For quantification of the localization data, we pooled the data points from individual intensity profiles for each dendrite segment and compared them statistically. Since this generates a relatively high number of data points, samples are quite likely to be significantly different. Acknowledging this, we now only compare the most distal segments (0-2 μm, 2-4 μm). We describe our approach now in more detail in the methods section.

Better description of fluorescence quantifications in Figs. 5, 7 in Methods section.

(2) Abstract "while plus-en out MT just grow from the soma" - I think this is probably only partially correct. There are various players important for organizing/nucleating these microtubules (e.g., Augmin for local MT nucleation).

Thank you for this comment. This was oversimplified and we rephrased the abstract.

(3) "We and others have previously shown that c4da neuron dendrite pruning depends on local microtubule disassembly in proximal dendrites (Herzmann et al, 2017)." Nice to add other references.

Thank you for this comment. We added additional references.

(4) Figure 2 EB imaging. Where was this performed along the dendrite? Close to the CB, this may reflect CB MT growth entering the dendrite.

Thank you for this comment. We usually perform EB1 imaging in primary or secondary dendrites -- now mentioned in the text (p. 7). The idea that anterograde comets in this region could reflect microtubules entering from the soma is interesting, especially as the fast anterograde comets upon combined shot knockdown and Mical overexpression behave so differently. We take up this idea in the Discussion - thanks again!

(5) Figure 3H-L I cannot see the magenta in the overlays. Separate channels seem to be needed. And where is the axon? Moreover, in the text it is stated that the I 3H fusion protein is exclusively in the axon. And control is not in 3M.

Thank you for this comment. Our lacZ antibodies are relatively weak, so the axons are not always visible. We therefore changed the text to only mention the soma now. We also now include all shown genotypes in the quantification.

(6) Figure 4A I'm not sure what is seen in the image. Is this a single dendrite? And if so, why is Shot only seen on one side?

Thank you for this comment. In this panel (now moved to Fig. 5A) we had to take a fairly thin projection of a confocal stack because of strong surrounding tissue signal. Because of this, the Shot::GFP signal on one edge of the dendrite is not as well visible. We try to improve this by indicating the edges of the dendrite in the panel.

Fig. 5A - D, indicated dendrite outline in confocal/SIM images.

(7) Figure S2C description is not the same as in the graph (neurons / dendrites).

Thank you for pointing this out. We corrected the legend (now Fig. 2C).

(8) Figure S4 - is this a single dendrite or a bundle?

The images (now Fig. S5) correspond to the microtubule signal from single primary dendrites, i. e., mostly microtubule bundles. We clarify now in the legend.

Fig. S5, specified nature of samples in legend.

Reviewer #2 (Recommendations for the authors):

(1) In the Discussion, page 13, the authors state that: "On the one hand, we provide evidence that Shot anchors microtubules via actin in mature neurons." This is not supported by the authors' evidence; indeed, the authors themselves state on page 10 that "... the CH1 domain is unlikely to be required for Shot cortical localization in dendritic shafts." If the authors mean to say that the cortical localization of Shot is restricted to dendrite tips, then superresolution microscopy should be done on dendrite tips.

Thank you for pointing this out. As the CH1 domain is clearly not required for Shot third instar localization, we agree with this notion and rephrase our interpretation accordingly: "Our data also do not rule out that Shot could anchor dendritic microtubules in mature third instar neurons, likely in cooperation with actin." (p. 17)

(2) The authors characterize the localization of Shot during larval stages but not during pupal stages. The authors should characterize where Shot localizes during dendrite pruning, since it is possible that Shot may localize differently at this stage. For example, the authors mention that Shot localizes to dendrite tips during early larval stages but not late larval stages; it seems that the localization of Shot is dynamic.

Thank you for this comment. In response, we analyzed Shot localization in a time-course experiment including the second instar larval stage and at 5 h APF in the early pupal stage (new Figure S4). This showed that Shot localizes almost exclusively to dendrite tips at the first instar, but is more broadly distributed in the soma and along dendrites shafts at all later developmental stages.

New Fig. S4, time course of Shot::GFP localization including first to third instars and early pupal stage.

(3) Figure 3: Khc-lacZ is not an ideal read-out of microtubule polarity per se. Better, read-out microtubule polarity using EB1-GFP.

Thank you for this comment. Unfortunately, this experiment is not feasible in this situation. In our MARCM system, we use a red fluorescent tdTomato to label c4da neurons, and all Shot transgenes carry GFP tags, prohibiting the use of both fluorescent EB1 transgenes that we have, EB1::GFP and EB1::mScarlet3.

(4) Figure 4 A-A' (also Figure S4): Superresolution microscopy data are presented but not analyzed. Analysis should be included before drawing a conclusion from these data.

Thank you for this comment. In response, we quantified the width of the dendritic microtubule bundles in the STED experiments (now Fig. S5) and found that these bundles are significantly thinner upon shot knockdown.

New Fig. S5 B, quantification of STED data.

(5) Figure 5, A and B: These images are extremely difficult to interpret on their own. It is not clear what the authors are interpreting as a positive signal. Analysis and quantification of Shot-FL and Shot-deltaCH1 should be included, not just representative images. This seems like a missed opportunity.

Thank you for this comment. In response, we show higher magnification images of Shot(endo)GFP, UAS-Shot::GFP, and the two actin-binding mutants (UAS-Shot ΔCH1::GFP and UAS-Shot ΔABD::GFP. For better interpretation, we indicate the boundaries of the dendrites as determined by tdTomato expression in the neurons (new Fig. 5 A - D).

(6) Figure 6: If gamma-tubulin is knocked down, does this decrease the number of EB1-GFP comets that originate from Rab11(+)-endosomes?

Thank you for this comment. This experiment is technically challenging because of the number and location of UAS transgenes involved. In addition, gamma-tubulin knockdown only has minor effects on pruning (no defects) or microtubule behavior (no orientation defects) in c4da neurons (Wang et al., 2019, eLife (Fengwei Yu lab).

(7) Figure 6: Rab11(+) endosomes correlate with the initiation of EB1-GFP comets in first instar larvae, but the orientation of microtubules is analyzed in third instar larvae -is there an effect on microtubule polarity in younger larvae? What about second instar larvae? On a related note, what is the orientation of microtubules that originate from Rab11(+) endosomes? These should be quantified data.

Thank you for this comment. We analyzed the effect of Rab11 knockdown at the first instar stage and found that this led to a non-significant increase in anterograde comets. These data are now included as new Fig. S6. We also checked the orientation and found that all comets from Rab11-cherry puncta were retrograde, this is now mentioned in the text.

New Fig. S6, effect of Rab11 dsRNA on comet orientation in first instar neurons.

(8) Figure 6: Panel G shows that Rab11-GFP and Msps-FLAG co-immunoprecipitate. What effect does Rab11[S25N] have on EB1-GFP comet frequency and site of origin?

Thank you for this comment. Unfortunately, Rab11[S25N] does not cause c4da neuron dendrite pruning phenotypes in our hands, even though it has worked in another lab (Lin et al., 2020, PLoS Genetics), possibly due to subtle differences in the GAL4 transgenes used. We therefore did not try this transgene in our assays.

(9) Figure 6: Does Patronin colocalize with Rab11 in neurons?

Thank you for this comment. We had originally included an image of Patronin::GFP expressed in c4da neurons that appeared to show tip enrichment. When we tried to reproduce and quantify this result, we found that Patronin localization varied strongly with expression levels, and more focused or punctate signals seemed to reflect either aggregation or decoration of microtubules. We therefore removed the image and instead included a more detailed genetic analysis of Patronin, which shows similar phenotypes as Shot (MT orientation defects at the first instar, enhancement of Shot knockdown phenotypes (Fig. 6).

Removed Patronin::GFP image (old Fig. 7H), added effect of Patronin dsRNA on comet orientation at first instar, enhancement of Shot dsRNA phenotype (new Fig. 6 D, E).

(10) Figure 7D: These data showing EB1-mScarlet3 comets originating from Shot-GFP should be quantified. Also, which Shot domains are involved in this recruitment?

Thank you for this comment. We observed only very few comets directly from tips in these experiments which precluded quantification. Acknowledging this, we moved the image to Supplementary Fig. 5.

As EB1 likely binds to Shot via SxIP motifs in the C-terminal region, we also tested whether the ShotΔCTail construct lacking the GAS2 domain and adjacent regions can recruit EB1. This is not the case. We therefore conclude that Shot recruits EB1 via its C-terminal part (new Figure 6D, E).

(11) Figure 2, A and B; Figure 4 B; Figure 6, B and D: Please include a scale bar for the time axis on these (and any other) kymographs.

Thank you for this comment. We checked and added time scale bars if missing.

(12) The authors should analyze the localization of GFP-tagged endogenous Shot in parallel to the exogenous expression of Shot under Gal4-UAS control.

Thank you for this comment. We tried to image endogenously tagged Shot::GFP at the first instar timepoint, but this was unfortunately not feasible due to very high Shot::GFP expression in neighboring tissues at this stage (and relatively low expression in c4da neurons).

(13) The authors should clarify at some point in the manuscript that EB1 comets can reflect either de novo microtubule growth (microtubule nucleation) or growth from the pre-existing ends of microtubules. EB1 comets alone do not indicate microtubule nucleation.

Thank you for this comment. We added the following sentence to the description of the first EB1 experiment (p. 7): "EB1 binds to the plus ends of growing microtubules (both newly nucleated and re-elongating), and plus end-bound EB1 is visible as moving dots (also known as comets)."

(14) It is unclear what the n in the graphs in the figures represent: neurons? Or dendrites/comets/etc? Please specify in the figure legend.

Thank you for this comment. We now specify this for each graph.

(15) Page 6, the authors write that "We and others have previously shown that c4da neuron dendrite pruning depends on local microtubule disassembly in proximal dendrites," but only cite their paper; please include the citations for the work by others.

Thank you for this comment. We added additional references.

(16) Page 9, the authors state: "Such phenomena are often seen when microtubules are not attached to specific anchoring sites such that they can be moved by microtubule motors." This is a fairly speculative interpretation of the data and should be saved for the Discussion.

Thank you for this comment. We moved this speculation to the Discussion.

Version 2

Abstract

The differential microtubule organization in axons and dendrites underlies neuronal polarity and developmental processes like neurite pruning. How neurons achieve their specific microtubule organization during development is an area of active research, and it has been especially hard to explain how dendritic microtubules are oriented with their plus ends towards the soma. Transient microtubule nucleation from tips of early growing dendrites has been detected in some systems and would explain how orientation is set up. In a survey for cytoskeletal regulators involved in dendrite pruning and microtubule organization in Drosophila, we found the spectraplakin Short stop (Shot), an actin/microtubule crosslinker. Loss of Shot causes microtubule orientation defects already during early dendrite development, when Shot is transiently recruited to tips of growing dendrites via its actin binding domain. Genetic and functional evidence suggest that Shot’s primary function in this process is to locally stabilize microtubules. We also provide evidence for a developmentally transient microtubule nucleation mechanism. Our data highlight the importance of transient and localized microtubule regulation for dendritic microtubule organization.

Introduction

Microtubules are long polar tube-like polymers composed of α/β-tubulin dimers that provide mechanical stability to cells and act as tracks for motor-based transport. Tubulin dimers are preferentially added, and also lost, at microtubule plus ends, whereas microtubule minus ends are relatively inert and often localize to microtubule organizing centers (MTOCs), structures of diverse composition that nucleate or anchor microtubules. In dividing cells, the main MTOCs are centrosomes, but many non-dividing or differentiated cells have less well characterized non-centrosomal (nc)MTOCs (Akhmanova & Kapitein, 2022).

Given their intrinsic polarity, microtubules, and the way they are organized in a given cell, play an important role in cellular polarity. This is particularly evident in neurons. In axons, all microtubules are oriented with their plus ends towards the distal synapses, thus enabling axon-specific transport, especially via kinesin-1. In contrast, a large fraction of dendritic microtubules is oriented with the plus ends towards the soma (“plus end-in”), enabling dendrite-specific transport via other motors including dynein. The plus end-in orientation is especially prevalent in invertebrates, where microtubules in primary dendrite branches are often almost uniformly plus end-in (Rolls et al, 2007). Despite the recognized importance of microtubule organization for neuronal function, it is still poorly understood how neuronal - and in particular, dendritic - microtubule organization is set up and maintained.

The differential organization of neuronal microtubules also underlies important neuronal morphogenetic processes. For example, the peripheral nociceptive class IV dendritic arborization (c4da) neurons of Drosophila larvae prune their long and branched sensory dendrites at the onset of metamorphosis. Pruning involves severing of proximal dendrite regions and subsequent degeneration of distal fragments (Rumpf et al, 2017; Furusawa & Emoto, 2021). The presumptive severing sites are marked by the local loss of microtubules, which predisposes them to mechanical tearing (Krämer et al, 2023). This local microtubule disassembly is initiated by an increase in microtubule dynamics at the onset of the pupal stage (Herzmann et al, 2017). Due to their uniform plus end-in orientation, the dendritic microtubules shrink from their plus ends in a coordinated fashion, eventually leading to the complete loss of microtubules in proximal dendrite regions (Herzmann et al, 2018; Rumpf et al, 2019). In support of this model, conditions that cause mixed orientation of dendritic microtubules either cause pruning defects or act as genetic enhancers of pruning mutants with microtubule-related functions. For example, the motor protein kinesin-2 together with its binding partner, the plus end-binder End-Binding protein 1 (EB1), guide polymerizing microtubules along preexisting ones, thus ensuring the maintenance of plus end-in orientation (Mattie et al, 2010). Loss of kinesin-2 causes pruning defects, while EB1 knockdown genetically enhances the pruning defects caused by the tau kinase Par-1 (Herzmann et al, 2018). Dendritic microtubule orientation and pruning defects are also caused by loss of the minus end binding protein Patronin (Wang et al, 2019) and the serine/threonine phosphatase PP2A (Rui et al, 2020; Wolterhoff et al, 2020). Loss of the microtubule polymerase Minispindles (Msps)/XMAP-215 (Tang et al, 2020) also causes defects in microtubule orientation, possibly linking dendritic microtubule orientation to microtubule growth or nucleation.

While several factors are known to maintain uniform dendritic microtubule orientation in Drosophila peripheral neurons, it is not clear how dendritic microtubule orientation arises during development. One model for this has been put forward based on the C. elegans PVD neuron model system. Here, a ncMTOC residing on Rab11-positive recycling endosomes is localized at the tip of the outgrowing dendrite, such that microtubules can only grow towards the soma, i. e., plus end-in (Liang et al, 2020; Harterink et al, 2018). Whether tip-based mechanisms could also inform dendritic microtubule polarity in other systems is not known.

Spectraplakins are a family of actin/microtubule cytoskeletal crosslinker proteins with important roles in microtubule organization in differentiated cells (Sanchez et al, 2021; Sun et al, 2019; Voelzmann et al, 2017). Anchored to the cell cortex or other actin-rich structures via an N-terminal actin binding domain, they can tether microtubules via binding sites for the minus end binding protein Patronin (Noordstra et al, 2016; Ning et al, 2016; Nashchekin et al, 2016), furthermore, spectraplakins can recruit factors related to microtubule growth such as the plus end binder EB1 (Slep et al, 2005). A neuronal function has been shown for the Drosophila spectraplakin Short stop (Shot), which bundles and stabilizes microtubules in axons (Alves-Silva et al, 2012, Okenve-Ramos et al, 2024). This effect is mediated by its C-terminal GAS2 domain (Okenve-Ramos et al, 2024), a type of microtubule-binding domains known to bundle and sometimes even nucleate microtubules (An et al, 2025).

Here, we describe an important role for Shot in dendritic microtubule organization. Loss of Shot in c4da neurons causes mixed orientation of dendritic microtubules and consequently, dendrite pruning defects. We find that Shot is already required for uniform orientation during early dendrite growth stages, when it transiently localizes to dendrite tips in an actin-dependent manner. Domain analyses show that this actin-dependent localization is required for Shot’s function. The GAS2 domain is also required, and Shot can increase dendritic microtubule density, indicating that it contributes to dendritic microtubule organization via a local stabilization mechanism. We also provide evidence that Rab11 is important for dendritic microtubule orientation and might be part of a transient dendritic MTOC. Our data suggest that tip-based mechanisms could be broadly involved in dendritic microtubule organization and identify Shot as an important player.

Results

Shot is required for c4da neuron dendrite pruning

The long and branched dendrites of third instar larval c4da neurons are completely pruned at the onset of the pupal stage, such that by 16 hours after puparium formation (h APF), they have lost all their dendrites (Fig. 1 A, 1 B, B’). In order to identify new regulators of c4da neuron dendrite pruning, we expressed candidate RNAis in c4da neurons under pickpocket (ppk)-GAL4 and assessed their effects at 16 h APF. This approach identified a dsRNA line targeting Shot, which encodes the sole Drosophila member of the spectraplakins, a family of large adaptor proteins with binding sites for actin and microtubules. C4da neurons expressing shot dsRNA had long and branched dendrites at the larval stage (Fig. 1 C), even though their dendritic field coverage seemed somewhat reduced compared to control neurons. At 16 h APF, dendrites could still be seen attached to the soma in almost half the c4da neurons expressing shot dsRNA (Fig. 1 C’, H - J), indicative of pruning defects. In order to confirm this result, we used Mosaic Analysis with a Repressible Cell Marker (MARCM) to generate c4da neuron clones homozygous for the loss-of-function allele shot3 (Kolodziej et al, 1995). shot3 mutant c4da neurons had shorter dendrites at the L3 stage compared to control MARCM c4da neuron clones (Fig. 1 D, E). At 16 h APF, most shot3 mutant neurons still had dendrites attached to the soma (Fig. 1 D’, E’, H - J). We also generated an sgRNA line targeting an exon common to all Shot splice isoforms and found that it caused significant c4da neuron dendrite pruning defects when expressed in c4da neurons in a sensitized shot3/+ heterozygous background (Fig. 1 F’, G’, H - J).

Shot is required for c4da neuron developmental dendrite pruning.

A Schematic showing the progression of c4da neuron dendrite pruning via hormonal induction (0 h APF) to severing (12 h APF) and clearance (16 h APF). B – G’ C4da neurons of the indicated genotypes were imaged at the third instar larval stage (B - G) and at 16 hours after puparium formation (h APF) (B’ - G’). B, B’ Control c4da neurons expressing Orco dsRNA under the control of ppk-GAL4. C, C’ C4da neurons expressing Shot dsRNA under the control of ppk-GAL4. D, D’ Control c4da neurons labeled by MARCM. E, E’ shot3mutant c4da neurons labeled by MARCM. F, F’ Control c4da neurons expressing Cas9P2 under ppk-GAL4 in a shot3 heterozygous background. G, G’ C4da neurons coexpressing Cas9P2 and Shot sgRNA under the control of ppk-GAL4 in a shot3 heterozygous background. H Phenotypic penetrance of pruning defects in B’ - G’. Number of neurons for each sample are indicated above the graph. *** P<0.001, **** P<0.0001, two-tailed Fisher’s exact test. I Average number of primary and secondary dendrites attached to the soma at 16 h APF in samples B’ - G’. Values are mean +/- s. e. m., **** P<0.0001, Mann-Whitney U test. J Total length of remaining dendrites at 16 h APF in samples B’ - G’. Values are mean +/- s. e. m., **** P<0.0001, Mann-Whitney U test. Scale bars in B and B’ are 50 µm.

C4da neurons undergo extensive remodeling during the pupal stage. Concomitantly with their dendrites, they also prune their presynapses in an ecdysone- and actin-dependent manner (Furusawa et al, 2023; Frommeyer et al, 2026). Later during the pupal stage, c4da neurons grow new dendrites adapted to the adult stage (Sanal et al, 2023). While Shot was not required for presynapse pruning (Figure S1 A), c4da neurons lacking Shot also failed to initiate dendrite regrowth at 72 h APF, when control c4da neurons had regrown long and branched dendrites (Figure S1 B, C). We conclude that Shot is required for c4da neuron dendrite remodeling, but does not seem to affect synapses.

Shot is required for uniform dendritic microtubule orientation

We and others have previously shown that c4da neuron dendrite pruning depends on local microtubule disassembly in proximal dendrites (Williams et al, 2005, Herzmann et al, 2017, Herzmann et al, 2018, Wang et al, 2019). To assess a potential effect of Shot knockdown on dendritic microtubule regulation during the early pupal stage, we visualized dendritic microtubules at 6 h APF using immunofluorescence against the microtubule-associated protein futsch/MAP1B. While control neurons consistently showed gaps in their futsch signal in these regions (Fig. 2 A, C), Shot knockdown neurons often still displayed continuous futsch staining in their proximal dendrites (Fig. 2 B, C).

Shot is required for uniform plus end-in orientation of dendritic microtubules.

A - C Delayed microtubule disassembly in neurons lacking Shot. Microtubules in control or Shot knockdown c4da neurons (green) were visualized by futsch/22C10 staining (magenta) at 6 h APF. Smaller panels below show futsch/22C10 signal in boxed areas in A, B, arrows indicate c4da neuron dendrites. A Control c4da neuron. B C4da neuron expressing Shot dsRNA. C Graph depicting the average number of dendrites with continuous futsch/22C10 staining per neuron in A, B. Box represents s. d., N = 10 neurons each, ** P<0.01, student’s t-test. D, E EB1::GFP comets were imaged in primary dendrites of third instar larval c4da neurons, and comet movement was depicted in kymographs. Time and direction of the soma are indicated by arrows. D Upper panel, EB1::GFP comets in control neuron expressing Orco dsRNA; lower panel, EB1::GFP comets in neuron expressing Shot dsRNA. E Upper panel, EB1::GFP comets in control c4da neuron MARCM clone; lower panel, EB1::GFP comets in shot3mutant c4da neuron MARCM clone. F Percentage of anterogradely moving comets in panels D, E. Numbers of comets per sample are indicated above the graph. ** P<0.01, **** P<0.0001, two-tailed Fisher’s exact test. G EB1::GFP comet speed in panels D, E. **** P<0.0001, Wilcoxon’s test. Scale bars in A and D are 20 and 5 µm, respectively.

As microtubules shrink away from the cell body in a coordinated fashion during dendrite pruning, the uniform plus end-in orientation of dendritic microtubules is particularly important for dendritic microtubule disassembly (Herzmann et al, 2018; Rumpf et al, 2019). To assess whether loss of Shot affects dendritic microtubule orientation, we expressed EB1::GFP in c4da neurons and live imaged it in dendrites. EB1 binds to the plus ends of growing microtubules (both newly nucleated and re-elongating), and plus end-bound EB1 is visible as moving dots (also known as comets). The movement direction of EB1 comets then indicates the orientation of the respective microtubule. As expected, EB1::GFP comets in primary or secondary dendrites of third instar control c4da neurons moved almost exclusively retrogradely towards the soma, indicating plus end-in orientation (Fig. 2 D, F). Upon shot knockdown, a significantly higher proportion of EB1::GFP comets could be seen moving anterogradely, or plus end-out (Fig. 2 D, F). To confirm this result, we assessed EB1::GFP behavior in shot3 c4da neuron MARCM clones and observed an even higher proportion of plus end-out dendritic microtubules in comparison to control c4da neurons MARCM clones (Fig. 2 E, F). Comets in shot3 mutant neurons also moved at a slightly higher speed than in control neurons (Fig. 2 G). We conclude that Shot is required for the correct orientation of dendritic microtubules.

Actin and microtubule binding by Shot are crucial for pruning and microtubule organization

In order to better understand the function of Shot in dendritic pruning and microtubule orientation, we next asked which domains of Shot are crucial for these functions. Of particular interest were the domains of Shot that bind to actin and microtubules. Shot has two Calponin Homology (CH) domains at the N-terminus that mediate binding to specific actin structures, and a microtubule-binding GAS2 domain in its C-terminal region (Fig. 3 A). We next expressed full length or truncated Shot variants in shot3 mutant c4da neurons and asked whether they could rescue the pruning and microtubule orientation defects. We used a genomic Shot BAC rescue construct with a C-terminal YFP-tag and a UAS construct containing GFP-tagged full length Shot UAS (ShotFL::GFP) as full length Shot variants. To address the role of the actin binding domain, we tested ShotΔCH1::GFP, a Shot variant lacking the first CH domain with decreased affinity for actin (Lee & Kolodziej, 2002), and ShotΔABD::GFP, which lacks both CH domains. To address the function of microtubule binding, we either used a full length Shot variant lacking the GAS2 domain (ShotΔCTail::GFP), or Shot Cterm::YFP, a C-terminal Shot fragment containing only the GAS2 domain (Fig. 3 A). Of the above constructs, we noticed that ShotΔCH1::GFP, but not the others, caused pruning defects when overexpressed under two copies of the c4da neuron driver ppk-GAL4 (Fig. S2).

Shot domains important for pruning and microtubule organization.

A Schematic of Shot domain structure and used UAS-Shot constructs. Right, summary of the phenotypic characterization below. CH, Calponin homology. B - J Ability of the indicated Shot UAS constructs to rescue the pruning defects of shot3 mutant c4da neuron MARCM clones at 16 h APF. B shot3 mutant c4da neuron. C shot3mutant c4da neuron rescued by a genomic Shot::YFP BAC construct. D shot3 mutant c4da neuron expressing full-length Shot::GFP. E shot3mutant c4da neuron expressing ShotΔCH1::GFP. F shot3mutant c4da neuron expressing ShotΔABD::GFP. G shot3 mutant c4da neuron expressing ShotΔCTail::GFP. H shot3 mutant c4da neuron expressing Shot C-term::YFP. I Penetrance of pruning defects. Number of neurons per sample is indicated above the graph. * P<0.05, **** P<0.0001, two-tailed Fisher’s exact test. J Number of primary and secondary dendrites attached to the cell body at the indicated timepoints. Values are mean +/- s. e. m., **** p<0.0001, Mann-Whitney U test. K – R Kin::lacZ localization in third instar shot3 c4da neuron MARCM clones. K shot3 mutant c4da neuron. L shot3 mutant c4da neuron rescued by a genomic Shot::YFP BAC construct. M shot3 mutant c4da neuron expressing full-length Shot::GFP. N shot3mutant c4da neuron expressing ShotΔCH1::GFP. O shot3mutant c4da neuron expressing ShotΔABD::GFP. P shot3 mutant c4da neuron expressing ShotΔCtail::GFP. Q shot3 mutant c4da neuron expressing Shot C-term::YFP. Asterisks in K – Q denote soma position, and arrows mark dendritic kin::lacZ puncta. R Number of dendritic kin::lacZ puncta in third instar c4da neurons expressing the indicated Shot UAS constructs. Number of neurons per sample is indicated above the graph. Values are mean +/- s. e. m., ** p<0.01, Mann-Whitney U test. Scale bars in B and K are 50 μm.

Both the genomic Shot BAC construct and UAS-ShotFL::GFP rescued the pruning defects of shot3 mutant neurons (Fig. 3 B - D, I, J). We next tested the domain mutants and found that while ShotΔCH1::GFP slightly reduced the pruning defects of the mutant (Fig. 3 E, I, J), ShotΔABD::GFP and ShotΔCTail::GFP could not rescue them at all (Fig. 3 F, G, I, J). Strikingly, expression of Shot Cterm::YFP containing only the GAS2 domain led to a full rescue of the pruning defects (Fig. 3 H - J).

We next asked whether these domains are also important for Shot’s function in dendritic microtubule orientation. As the GFP tags of the Shot transgenes and the tdtomato fluorophore used to label MARCM clones precluded the use of fluorescently tagged EB1 constructs, we used kinesin-lacZ (kin::lacZ), a fusion between the kinesin-1 motor domain and β-galactosidase. Because the kinesin motor domain migrates towards microtubule plus ends, this fusion protein will always moves towards the soma in dendrites with normal plus end-in orientation and will thus show staining in the soma, but not dendrites. Consistent with the observation that loss of Shot causes microtubule orientation defects, kin::lacZ was depleted from the soma and found in dendritic puncta in shot3 mutant c4da neurons (Fig. 3 K, R), but in animals carrying the genomic Shot BAC, it was localized in the soma (Fig. 3 L, R). UAS-based expression of ShotFL::GFP in mutant neurons did not fully rescue kin::lacZ localization but reduced the number of dendritic kin::lacZ puncta (Fig. 3 M, R). shot3mutant c4da neurons expressing ShotΔCH1::GFP, ShotΔABD::GFP or ShotΔCTail::GFP did not confer rescue effects and showed many dendritic kin::lacZ puncta comparable to shot3 mutant neurons (Fig. 3 N - P, R). Again, expression of Shot Cterm::YFP restored dendritic exclusion of kin::lacZ, and it could be seen in the soma (Fig. 3 Q, R).

Thus, in the context of full length Shot, both its actin and microtubule binding activities are required for its roles in dendrite pruning and microtubule orientation, but the isolated GAS2 domain alone was sufficient to rescue Shot function. This is reminiscent of a previous study reporting that full length Shot is in an autoinhibited state that can be relieved by actin binding (Applewhite et al, 2013). In support of the notion that microtubule orientation defects cause pruning defects, As the pruning and microtubule orientation defects of shot mutant neurons go hand in hand in our domain analysis.

Shot cooperates with actin to promote microtubule orientation

To independently probe for a role for actin in dendritic microtubule organization, we sought to manipulate actin dynamics in vivo. As treatment with actin-depolymerizing drugs like latrunculin would likely be highly toxic to larvae, we considered Mical, a potent actin severing enzyme that breaks actin filaments by oxidizing actin on specific methionine residues (Grintsevich et al, 2017). Mical is highly specific for actin (Rajan et al, 2023) and its overexpression can disrupt actin bundles (Hung et al, 2010). Mical overexpression in larval c4da neurons did not disrupt dendrite morphology, but altered dendritic F-actin distribution at the first instar larval stage, as assessed by lifeact::GFP (Fig. S3). We next tested whether Mical overexpression has an effect on dendritic microtubule orientation and found that it caused a significantly increased number of EB1::GFP comets to move anterogradely (Fig. 4 A, B). When combined with Shot knockdown, Mical overexpression did not lead to an additive increase in the percentage of anterograde comets, potentially indicating epistasis between the manipulations (Fig. 4 A, B). However, we noted that the anterograde comets in the combined Shot knockdown/Mical overexpression neurons moved significantly faster than the retrograde comets, and also faster than the anterograde comets in the single manipulations (Fig. 4 C). Movement of the anterograde comets in these neurons was also more irregular than in other samples, as these comets frequently slowed down and then sped up again (Fig. 4 A).

Shot and actin cooperatively regulate the behavior of dendritic microtubules.

A Effect of actin severing on EB1::GFP comet movement. Mical was overexpressed in c4da neurons with or without Shot knockdown, and EB1 comets were analyzed as in Fig. 2 B. Upper panel, kymograph of EB1 comet movement in c4da neuron overexpressing Mical. Lower panel, kymograph of EB1 comet movement upon Mical overexpression and shot knockdown. B Penetrance of anterograde dendritic EB1 comets in control c4da neuron expressing orco dsRNA, or in neurons expressing shot dsRNA, Mical, or both (kymographs for the latter shown in A). Number of comets per sample is given in the graph. * P<0.05, Fisher’s exact test. C Speed of dendritic EB1 comets in A, B. **** P<0.0001, Wilcoxon’s test. D - F Synergistic effects of Shot knockdown and Mical overexpression on c4da neuron dendrite pruning. D C4da neuron expressing shot dsRNA and a control UAS construct at 16 h APF. E C4da neuron expressing shot dsRNA and UAS-Mical at 16 h APF. F Number of primary and secondary dendrites attached to the soma in c4da neurons overexpressing Mical, upon Shot knockdown, or both at 16 h APF. Number of neurons per sample is given in the graph. *** P<0.001, Mann-Whitney U test. Scale bars are 5 μm in A and 50 μm in B.

Conditions that cause mixed dendritic microtubule orientation also either cause c4da neuron dendrite pruning defects or enhance them (Herzmann et al., 2018). Mical overexpression alone does not cause pruning defects - in fact, it can suppress the pruning defects of certain ecdysone signaling mutants (Kirilly et al, 2009; Rode et al, 2018). However, we observed a synergistic enhancement of the pruning defects caused by Shot knockdown when we overexpressed Mical (Fig. 4 D - F). Thus, our data suggest that Shot and actin cooperatively regulate dendritic microtubule orientation.

Shot is recruited to dendrite tips during early dendrite growth

As Shot is often recruited to specific subcellular locations via its actin binding domain (Nashchekin et al, 2024), we next assessed Shot localization in dendrites using GFP-tagged Shot constructs. We first visualized endogenously GFP-tagged Shot in third instar c4da neurons using structured illumination microscopy (SIM). In dendrite shafts, endogenously GFP-tagged Shot localized in closely spaced puncta close to the plasma membrane (Fig. 5 A), and transgenically expressed full length Shot (ShotFL::GFP) localized in a similar pattern (Fig. 5 B). We then visualized the two Shot variants lacking high affinity actin binding, ShotΔCH1::GFP and ShotΔABD::GFP, and found that they also localized close to the plasma membrane in a very similar pattern to full-length Shot (Fig. 5 C, D). Thus, actin binding is unlikely to be required for Shot cortical localization in dendritic shafts.

Shot is recruited to tips of growing dendrites via its actin binding domain.

The indicated GFP-tagged Shot constructs and tdtomato were expressed in c4da neurons under ppk-GAL4 and visualized at the indicated developmental stages. A - D Actin binding is not required for Shot localization in shafts of mature dendrites. The indicated Shot::GFP constructs were visualized in third instar primary c4da neuron dendrites by immunofluorescence followed by structured illumination microscopy. Borders of dendrites are indicated by white broken lines. A Endogenously tagged Shot::GFP. B Localization of transgenic full-length Shot (ShotFL::GFP) expressed under ppk-GAL4. C Localization of ShotΔCH1::GFP expressed under ppk-GAL4. D Localization of ShotΔABD::GFP expressed under ppk-GAL4. E - K Localization of the indicated Shot::GFP variants in first instar c4da neurons. Images on the right show close-ups of dendrite tip (1) and soma regions (2). Asterisks in the dendrite close-ups mark the position of the tip. E Localization of ShotFL::GFP. F Localization of ShotΔCH1::GFP. G Localization of ShotΔABD::GFP. H Fluorescence intensity profiles of Shot::GFP variants in the distal 10 μm of first instar dendrites in E - G. Intensity values are normalized to Shot::GFP intensity in the cell body. Solid lines indicate average, envelopes indicate s. d. (N=10 neurons). Compared to full length Shot, GFP intensities of ShotΔCH1 and ShotΔABD in the 0-2 μm and 2-4 μm brackets from the tip were significantly reduced (P<0.0001, Mann Whitney U test). I Dendrite tip occupancy of Shot::GFP variants in E - G. ** P<0.01, Fisher’s exact test. J, K Effect of Mical overexpression on ShotFL::GFP localization. J Localization of ShotFL::GFP in a c4da neuron overexpressing Mical. K Fluorescence intensity profiles of ShotFL::GFP in the distal 10 μm of first instar dendrites with or without Mical overexpression. Shot::GFP intensity upon Mical overexpression was not significantly different from Shot::GFP without Mical in the 0-2 μm bracket, but highly significant in the 2-4 μm bracket (N=10 neurons, P<0.0001, Mann Whitney U test). Scale bars are 5 μm in A and 10 μm in E (larger image and close-up).

The control of dendritic microtubule organization can likely be divided in early establishment and later maintenance processes. To assess potential Shot functions at other developmental stages, we investigated the localization of transgenic Shot::GFP from the first larval instar until the early pupal stage (Fig. S4). Due to strong background from neighboring tissues, we were not able to visualize endogenously GFP-tagged Shot at all stages and throughout the whole c4da neuron. Strikingly, Shot::GFP was highly enriched at dendrite tips in first instar c4da neurons, but could only be detected at low levels in the soma and proximal dendrites (Fig. 5 E, H, I). From the second instar onwards, Shot::GFP appeared relatively evenly distributed in the soma and along both axons and dendrites (Fig. S4), indicating that Shot dynamically changes its localization during development. To test whether Shot dendrite tip localization depends on actin, we next imaged the actin binding-deficient variants ShotΔCH1::GFP and ShotΔABD::GFP in first instar c4da neurons. ShotΔCH1::GFP accumulated in several dots per dendrite that could be many micrometers away from the tips (Fig. 5 F, H). Overall, significantly fewer distal dendrites were occupied by ShotΔCH1::GFP compared to full length Shot (Fig. 5 F, I). ShotΔABD::GFP was found at tips (Fig. 5 G, I), but always also localized in dots along dendritic shafts, such that there was no effective concentration at tips (Fig. 5 G, H). To independently test whether Shot tip localization is actin-dependent, we used UAS-Mical to disrupt actin. In c4da neurons overexpressing Mical, the Shot signal at tips was significantly broadened and Shot puncta could be seen along dendrite shafts (Fig. 5 J, K), similar to what was observed with ShotΔABD::GFP. We conclude that actin binding is required to transiently recruit and concentrate Shot at dendrite tips during the early stages of c4da neuron dendrite growth.

Shot acts early and is linked to end binding proteins

The observation that the Shot actin binding domain is required for both early dendrite tip localization and uniform dendritic microtubule orientation suggests that Shot function during early dendrite development is crucial for microtubule polarity. We therefore tested whether loss of Shot already causes defects in microtubule orientation during early development. When we imaged the directionality of EB1::GFP comets in first instar c4da neuron dendrites, we indeed found a significant increase in plus end-out microtubules upon loss of Shot (Fig. 6 A).

Shot acts during early development and interacts with end binding proteins.

A Frequency of anterograde EB1 comets at the L1 stage in control (Orco dsRNA) and shot knockdown neurons (shot dsRNA, shot3/+). Number of comets per sample is shown in the graph, * P<0.05, Mann Whitney U test. B Microtubules in dendrites of first instar control neurons or neurons expressing Shot::GFP were visualized using mCherry-tagged α-tubulin in single confocal slices. C Dendritic microtubule density for samples in B. Dendritic mCherry-α-tubulin 84B fluorescence intensity was normalized to that of the soma. N= 14 - 22 dendrites from 9-14 neurons, * P<0.05, Student’s t-test. D Shot can recruit EB1 to dendrite tips. EB1::mScarlet3 was expressed in c4da neurons alone, with ShotFL::GFP, or with ShotΔCTail::GFP, and visualized at the first instar larval stage. Positions of dendrite tips are indicated by asterisks. E Fluorescence intensity profiles of EB1::mScarlet3 in the distal 10 μm of first instar dendrites. Solid lines indicate average, envelopes indicate s. e. m. (N=9-10 dendrites, 3 neurons each). Significance between genotypes was calculated for 0 - 2 and 2 - 4 μm distances from the tip. **** P<0.0001, Mann Whitney U test. F Frequency of anterograde EB1 comets at the L1 stage in control (Orco dsRNA) and patronin knockdown neurons (patronin dsRNA). Number of comets is shown in the graph. ** P<0.01, Mann Whitney U test. G Frequency of anterograde EB1 comets at the L3 stage in shot knockdown neurons (shot dsRNA) and shot knockdown neurons in a heterozygous patronin mutant background. Number of comets is shown in the graph. * P<0.05, Mann Whitney U test. H - K Synergistic genetic interactions between shot and end binding proteins during c4da neuron dendrite pruning. H - J shot dsRNA was expressed in c4da neurons of animals in the indicated backgrounds, and pruning defects were quantified at 16 h APF. H C4da neuron expressing shot dsRNA. I C4da neuron expressing shot dsRNA in a patronin/+ background. J C4da neuron co-expressing shot and EB1 dsRNAs. I Severity of pruning defects in H - J and in respective controls. N = 33 - 52 neurons, ** P<0.01, **** P<0.0001, Wilcoxon’s test. Scale bars are 5 μm in B, 2 μm in D and 50 μm in E.

To assess how Shot might affect microtubules at this stage, we next imaged mCherry-tagged α-tubulin live in first instar neurons and manipulated Shot levels. Interestingly, Shot overexpression caused a clear increase in microtubule density in first instar dendrites (Fig. 6 B, C). Shot knockdown did not cause a significant decrease of the dendritic microtubule signal in this experiment (Fig. 6 C), but this could in part be due to technical reasons, as we could not use the strongest Shot manipulation (Shot knockdown only, instead of knockdown in a heterozygous shot mutant background). To visualize dendritic microtubule structure at high resolution in third instar larval c4da neurons with the stronger manipulation, we also used stimulated emission depletion (STED) microscopy (Fig. S5). While microtubules appeared as several continuous and bundled filaments in controls, fewer filamentous structures were visible in dendrites of c4da neurons lacking Shot, the microtubule staining was more punctate, and the width of the microtubule bundles significantly thinner. Together with the domain analysis demonstrating the importance of the GAS2 domain, these data suggest that Shot promotes stabilization of dendritic microtubules. Shot is known to bind to the plus end binding protein EB1 via several SxIP motifs in its C-terminal region (Slep et al, 2005; Alves-Silva et al, 2012), and EB1 is also implicated in dendritic microtubule orientation (Mattie et al., 2010). In the presence of Shot::GFP, red fluorescently tagged EB1 also became recruited to dendrite tips at the first instar and colocalized with Shot::GFP (Fig. 6 D, E). This depended on the presence of the Shot C-terminal region including the SxIP motifs, as ShotΔCTail expression did not recruit EB1 (Fig. 6 D, E). Very rarely, EB1 comets could be seen emanating from a Shot/EB1-rich tip (Fig. S5).

The minus end binder Patronin also interacts with Shot (Nashchekin et al, 2016) and is required for uniform plus end-in dendritic microtubule orientation (Wang et al., 2019). We confirmed in immunoprecipitation experiments that a Shot fragment comprising spectrin repeats 27-32 interacts with Patronin (Fig. S5), similar to what has been reported for the mammalian Shot and Patronin homologues (Noordstra et al, 2016; Ning et al, 2016). Similarly to Shot, the mechanism of Patronin in dendritic microtubule orientation likely involves microtubule stabilization (by protecting minus ends from deolymerization) (Wang et al., 2019). Patronin knockdown caused an increase in the frequency of plus end-out microtubules already at the first instar stage (Fig. 6 F), and a heterozygous P element mutant in Patronin significantly increased the percentage of plus end-out microtubules at the third instar when combined with Shot knockdown (Fig. 6 G). We also tested for genetic interactions between Shot and end binding proteins using pruning as readout and found that both the heterozygous Patronin mutant and a dsRNA construct targeting EB1 strongly enhanced the pruning phenotypes seen upon shot knockdown (Fig. 6 H - K), again demonstrating the link between dendritic microtubule orientation and pruning. Thus, Shot likely cooperates closely with end binders in dendrites.

Rab11 is linked to dendritic microtubule regulation in c4da neurons

A tip-based mechanism for microtubule orientation has first been proposed in C. elegans PVD sensory neurons, where Rab11-positive vesicles are part of a transient ncMTOC that localizes to the tips of growing dendrites (Liang et al, 2020; Harterink et al, 2018). We therefore asked whether a similar transient MTOC would also exist in Drosophila sensory neurons. In Drosophila sensory neurons, vesicle-based dendritic MTOCs have been proposed on either Golgi outposts (Ori-McKenney et al, 2012; Mukherjee et al, 2020; Yalgin et al, 2015) or Rab5 endosomes (Nguyen et al, 2014; Weiner et al, 2020), and microtubules have been observed originating in dendritic shafts, at branchpoints and at tips (Mukherjee et al, 2020; Yalgin et al, 2015; Weiner et al, 2020; Liang et al, 2020).

We and others have shown that Rab11 is required for c4da neuron dendrite pruning via an unclear mechanism (Krämer et al, 2019; Lin et al, 2020). To test whether Drosophila Rab11 interacts with microtubule regulators involved in pruning regulation, we used immunoprecipitation of tagged proteins from transfected S2 cells. This showed that Rab11 could be co-precipitated with Msps/XMAP215, a microtubule polymerase linked to pruning and dendritic microtubule organization (Tang et al, 2020) (Fig. 7 A). This interaction depended on GTP binding by Rab11, as GTP-binding-deficient Rab11 S25N showed a reduced interaction with Msps (Fig. 7 A). We next asked whether Rab11 colocalizes with potential MTOC components in c4da neuron dendrites using fluorescently tagged transgenes. At the first instar, Rab11::mCherry puncta could be seen overlapping with γ-tubulin::GFP both at dendrite branchpoints and tips (Fig. 7 B). To test whether such puncta could nucleate microtubules, we observed Rab11::mCherry together with EB1::GFP. In first instar neurons, a fraction of EB1::GFP comets could be observed originating at dendritic Rab11::mCherry puncta (6/9 neurons, 10/58 comets) in dendritic shafts or at branchpoints (Fig. 7 C, D). All of these comets moved retrogradely (10/10). Partially because of low EB1::GFP fluorescence, we could not detect dendrite tips - or comets originating from tips - in these experiments. In support of the idea that this Rab11/microtubule link might be transient, no EB1::GFP comets emanating from Rab11::mCherry puncta could be observed at the third instar stage (0/9 neurons, 27 comets) (Fig. 7 D). We next knocked down Rab11 and Rab5 and assessed the effects on microtubule orientation. Knockdown of Rab11, but not Rab5, caused a significant increase in the fraction of anterograde EB1 comets (Fig. 7 E, F) as well as increased comet speed at the third instar stage (Fig. 7 G). At the first instar stage, Rab11 knockdown caused a milder increase in anterograde comets that was not significant (Fig. S6). Thus, our evidence links Rab11 to dendritic microtubule regulation in Drosophila c4da neurons, but in part for technical reasons we cannot be certain whether it acts through the same tip-based mechanism as in C. elegans PVD.

Evidence that Rab11 is linked to developmental dendritic microtubule regulation.

A GFP-tagged Rab11 (wt or S25N) and FLAG-tagged Msps were cotransfected into S2 cells and immunoprecipitated with FLAG beads. Inputs and immunoprecipitates (IP) were blotted with the indicated antibodies. Sizes of molecular weight markers in kiloDalton (kD) are shown. IgG denotes antibody heavy chains. B Rab11 colocalizes with the MTOC marker γ-tubulin in growing c4da neuron dendrites. Rab11::mCherry and γ-tubulin23C::GFP were coexpressed under ppk-GAL4 and visualized at the first instar. Arrows show Rab11/ γ-tubulin-positive puncta at dendrite tips, the asterisk denotes a double-positive dot at a branchpoint. C Microtubules can originate at Rab11 puncta in dendrites. Rab11::mCherry and EB1::GFP were coexpressed in c4da neurons, and EB1 comets were visualized at the first instar. Example kymographs show Rab11::mCherry (magenta) and EB1::GFP (green). D Percentage of EB1 comets arising from dendritic Rab11 puncta at the first and third instars. Number of comets is shown in the graph. * P<0.05, Fisher’s exact test. E Effect of Rab11 knockdown on dendritic microtubule orientation. Kymographs show EB1::GFP movement in third instar c4da neuron dendrites. Upper panel, control c4da neuron expressing Orco dsRNA; middle panel, c4da neuron expressing Rab11 dsRNA; lower panel, c4da neuron expressing Rab5 dsRNA. F Penetrance of anterograde comets in D. Number of comets in E and F is shown in the graph. ** P<0.01, two-tailed Fisher’s exact’s test. G Speed of EB1::GFP comets in D. **** P<0.0001, Wilcoxon’s test. H - J Rab11 and Shot interact genetically during dendrite pruning. Rab11 dsRNA was expressed in c4da neurons of control animals (H) or in a shot3/+ heterozygous background (I), and neurons were visualized at 16 h APF. J Severity of pruning defects in H, I. N = 41 and 42 neurons. **** P<0.0001, Wilcoxon’s test. K Kymograph analysis of Rab11::GFP movement in dendrites of first instar control c4da neurons or shot knockdown neurons (lower panel). L Fraction of mobile Rab11::GFP particles in samples in K. Number of Rab11::GFP particles is shown in the graph. n. s., not significant, Fisher’s exact test. Scale bars are 5 μm in B, C, E, and K, and 20 μm in H.

As Shot has previously been shown to regulate Rab11 localization in epithelia (Khanal et al, 2016), we tested whether the two factors might also act in a common pathway in dendrites. We first used pruning phenotypic strength to assess a potential genetic interaction and found that Rab11 knockdown in a shot3/+ heterozygous background caused significantly more higher order dendrites to stay attached to the soma than with Rab11 knockdown alone, indicative of enhancement (Fig. 7 H - J). To observe potential effects in dendrites more directly, we expressed GFP-tagged Rab11 and visualized it using live microscopy at the first instar stage. Rab11::GFP vesicles in dendrites could be found along the whole dendritic shaft and were highly mobile (Fig. 7K). However, neither the mobility nor the overall localization of these puncta were changed upon shot knockdown (Fig. 7L). Thus, Shot is unlikely to directly regulate Rab11 trafficking in dendrites.

Discussion

Microtubule orientation in neurites is crucial for neuronal polarity and developmental processes like neurite pruning, but the mechanisms that shape neuronal microtubule organization during development are still emerging. Here, we show that the spectraplakin Shot is required for the uniform plus end-in organization of dendritic microtubules. Our data suggest that Shot acts in a developmental timing-dependent fashion via localized microtubule stabilization at dendrite tips. Shot transiently localizes to dendrite tips in early growing dendrites. As this localization is actin-dependent, Shot recruitment may be mediated via a transient actin-rich growth cone-like structure. Importantly, the actin-binding domain required for tip recruitment is also required for correct microtubule orientation, indicating that Shot via a tip-based mechanism to orient microtubules. The main function of Shot at dendrite tips appears to be microtubule stabilization via the GAS2 domain. The surprising finding that this domain is sufficient to rescue Shot function in the absence of the large spectrin repeat region is consistent with the idea that Shot is regulated via autoinhibition, and actin binding may be required to disinhibit it (Applewhite et al, 2013). In addition to direct microtubule stabilization via the GAS2 domain, Shot can recruit the plus end binder EB1 (and potentially Patronin) to increase microtubule stability. This tip-based mechanism is reminiscent of the mechanism in the C. elegans PVD neuron, where a transient MTOC exists at tips of early growing dendrites. It was recently shown that in addition to bundling and stabilization, GAS2 domains can also nucleate microtubules in vitro (An et al, 2025). We think that it is unlikely that Shot itself acts to nucleate microtubules at dendrite tips because we were not able to detect a clear increase in microtubule nucleation from tips upon Shot overexpression. However, it is interesting to speculate that selective local microtubule stabilization at tips may already be sufficient to favour plus end-in polarity if the microtubules generated in other regions are sufficiently unstable, and microtubules at tips can only grow in the retrograde direction. This idea could be investigated in the future, e. g., through mathematical modelling. However, it is also possible that other, additional MTOC components localize to early dendrite tips in Drosophila sensory neurons as well. One candidate here is Rab11 which our analysis links to dendritic microtubule regulation, even though we did not see Rab11 enriched at tips. It is possible that the first instar stage, when most of our reagents start to work reliably, is already too late to visualize some of the very early growth events. In any case, our data suggest that early-stage tip-based mechanisms are a general feature of dendritic development.

Our data also do not rule out that Shot could anchor dendritic microtubules in mature third instar neurons, likely in cooperation with actin. Loss of Shot in combination with actin disassembly cause erratic comet behaviors (Fig. 4), which could be indicative of sliding, e. g., through defective microtubule anchoring (He et al, 2020). Alternatively, the observation that only the plus end-out oriented subset of comets is fast-moving could point to a different origin of these microtubules, e. g., they could enter the dendrites from the soma. Local membrane-bound microtubule stabilizers have been linked to the maintenance of dendritic microtubule orientation (He et al, 2020, Liang et al, 2025), and Shot might have a similar function. Crucially, both proposed functions of Shot in dendritic microtubule organization depend on actin binding, linking the two main cytoskeletal components in neurite organization.

Last but not least, our data again confirm our notion that uniform plus end-in dendritic microtubule orientation is required for c4da neuron dendrite pruning, as all Shot manipulations leading to a more mixed orientation caused pruning defects. As we do not have evidence for Shot regulation at the onset of the pupal stage (Fig. S4), the function of Shot crucial for pruning is most arguably its early role in setting up uniform microtubule orientation.

Materials and Methods

Fly Strains

C4da neurons were labeled by UAS-CD8::GFP or UAS-tdtomato expression under ppk-GAL4 (Grueber et al, 2007). MARCM clones of shot3mutants (BL 5141) were induced with SOP-FLP (Matsubara et al, 2011) and labeled by tdtomato expression under nsyb-GAL4R57C10. Endogenously tagged Shot::GFP was from (Voelzmann et al, 2026) and the BAC carrying the SHOT genomic locus with a YFP tag was from (Nashchekin et al, 2016). UAS transgenes were UAS-ShotRE::GFP on the X chromosome (BL 29044), or mobilized to the third chromosome (this study), UAS-ShotRC::GFP (ΔCH1, BL 29042), UAS-ShotΔABD::GFP (Hahn et al, 2021), UAS-ShotΔCTail::GFP (Alves-Silva et al, 2012), UASp-Shot Cterm::YFP (Nashchekin et al, 2024), UAS-kin::lacZ (Clark et al, 1997), UAS-Mical (Terman et al, 2002), UAS-αtub84B::mCherry (Villars et al, 2022), UAS-Rab11::GFP (BL 8506), UAS-γtub23C::GFP (Nguyen et al, 2011), and UAS-cytoABKAR TA (Marzano et al, 2021) as inert UAS control, UAS-Brpshort::Strawberry to label presynapses (gift from S. Sigrist). EB1 analyses were carried out using either UAS-EB1::GFP (Zheng et al, 2008), ppk-EB1::GFP (Arthur et al, 2015), UAS-EB1::mScarlet3 (gift from M. Rolls) or ppk-EB1::mScarlet3 (this study). The following dsRNA lines were used: Orco (BL 31278) as control, Shot (BL 28336), Rab11 (VDRC 22198), Rab5 (VDRC 34096), EB1 (VDRC 24451). All dsRNAs were coexpressed with UAS-dcr2 (Dietzl et al, 2007). Additional mutant alleles were patronink07433 (Kyoto 111217).

Cloning and transgenes

Shot sgRNAs GCTGCCCTCTCAGGCCGATT (target 1) and GAGTTCTCCAGAGTGGTCAC (target 2) were cloned into pCFD4w+ (gift from S. Schirmeier). For transgenesis, plasmids were injected into flies carrying the 86Fb acceptor site. For biochemical experiments, the Rab11 ORF was cloned into pUAST attB carrying a C-terminal GFP tag using TOPO/TA cloning, The S25N mutation was inserted by PCR mutatgenesis. N-terminally GFP-tagged Patronin was cloned into pUAST using standard restriciton sites. To express the Shot spectrin repeats 27-32, a fragment of the Shot ΟRF encoding amino acids 3626 - 4248 was cloned into pUAST carrying a C-terminal myc tag via TOPO/TA cloning. For Rab11::mCherry flies, the Rab11 entry clone was used to generate pUAST attB Rab11::mCherry, the insert was subcloned into pUAST, which was then injected into w1118 flies using classical transposase-mediated transgenesis. To generate EB1::mScarlet3, a Drosophila codon optimized mScarlet3 (Gadella et al, 2023, synthesized by Twist Bioscience) was added after the EB1 coding sequence separated by a CACACCTCCACTACCGCTAGCAGGCCGGCCACGCGTGGTACCTTCTGGTCCA linker and cloned into pUAST. The resulting plasmid was injected into wild type embryos by Bestgene. The same insert was also inserted into the pDEST-APHIH vector carrying a ppk promotor fragment and injected into attP2 flies to generate ppk-EB1::mScarlet3 flies. C-terminally FLAG-tagged Msps (isoform RB) was cloned into pUAST attB using standard cloning procedures.

Immunofluorescence

Briefly, appropriately staged pupal filets were fixed in 4% formaldehyde, blocked in PBS with 0.3 % Triton X-100 and 10% goat serum, and incubated with antibody in blocking buffer over night. C4da neurons labeled by ppk promotor fusions, endogenous or transgenic GFP-tagged Shot were visualized with chicken (1:500, Aves labs), rabbit (1:1000, Invitrogen A11122) or mouse (1:1000, Invitrogen A1120) anti-GFP. Tdtomato and mcherry-tagged transgenes were detected with rat anti-mCherry (1:1000, Invitrogen) or rabbit anti-DsRed antibodies (1:1000, Clontech). Kin::β-galactosidase fusion proteins were detected with rabbit anti-β-galactosidase (Cappel, preabsorbed in-house, 1:700), and Drosophila futsch was detected with 22C10 (DSHB, 1:50). Secondary antibodies for regular immunofluorescence were conjugated to Alexa Fluor 488, 568 or 647. For STED microscopy of tagged tubulin, Abberior anti rabbit STAR RED were used as secondary antibodies.

Microscopy, time lapse imaging

For analyses of dendrite pruning phenotypes at 16 h APF, animals were dissected out of the pupal case and dorsal ddaC c4da neurons in segments A2 - A5 were imaged live on a Zeiss LSM710 confocal microscope with a 20x Plan Apochromat water objective (1.0 NA). For regrowth analyses at 72 h APF, animals were dissected out of the pupal case and ventral v’ada c4da neurons in segments A2 - A4 were imaged as described (Sanal et al, 2023). For live protein localization analyses, first instar larvae were immobilized on double sided tape, third instar larvae were briefly anaesthesized with ether before imaging. Neurons were then imaged live on Zeiss LSM710 or LSM880 microscopes with 63x Plan Apochromat DIC M27 oil objectives (1.4 NA). For EB1::GFP or EB1::mScarlet3 live imaging, third instar larvae were anaesthesized with ether, and first instar larvae were immobilized on double-sided tape. Live analysis was performed on a Zeiss LSM880 microscope using a 40x Plan Apochromat FCS M27 (1.2 NA) oil objective. Consecutive images of a single plane were taken every second for 1 - 2 minutes, Z positions were adjusted if larvae moved. Movies were stabilized with the Fiji Image Stabilizer plugin, and the KymoResliceWide plugin was used to generate EB1 comet kymographs. Comet speed was read out manually from kymographs.

Immunofluorescence images of kin::lacZ were obtained on a Zeiss LSM 710 microscope using a 40x Zeiss C Apochromat water objective (1.1 NA). Structured illumination microscopy (SIM) was performed on a Zeiss Elyra 7 microscope with a 63x Plan Apochromat DIC M27 oil objective (1.4 NA) with an Edge 4.2 SIM camera (50 ms exposure time), and deconvolution was done using SIM2 software. All microscopes used ZEN Black software. Most Stimulated Emission Depletion Microscopy (STED) images were taken with an abberior FACILITY LINE (Abberior Instruments GmbH) equipped with an inverted IX83 microscope (Olympus), a 60x oil objective (Olympus UPLXAPO 60XO, NA 1.42), using pulsed excitation lasers at 640 nm and a pulsed STED laser operating at 775 nm and gated detection with avalanche photodiode element (APD) detectors, where acquisition operations were controlled by Lightbox Software (Abberior Instruments GmbH). Some 2D STED images were also taken with an abberior STEDYCON with an inverted IX83 microscope (Olympus), a 100x oil objective (Olympus UPLXAPO100XO, NA 1.45), using pulsed excitation lasers at 640 nm and a pulsed STED laser operating at 775 nm, continuous autofocus and gated detection with avalanche photodiode element detectors, where the acquisition operations were controlled by STEDYCON Software (abberior Instruments), and images were deconvolved using Huygens deconvolution software.

Microscopic images shown are maximum projections or (where indicated) single plane images. All processing and measurements were done in Fiji (Schindelin et al, 2012) using the plug-ins NeuronJ for dendrite length measurements.

S2 cell culture, immunoprecipitation and Western blots

Appropriate pUAST plasmids (Rab11::GFP (wt or S25N), MspsFLAG, GFP::Patronin, myc::SRR27-32) were cotransfected into S2 cells with Actin5C-GAL4 as described (Herzmann et al, 2017). After 72 hours, cells were harvested in ice-cold PBS, and lysed in cold lysis buffer (100 mM NaCl, 5 mM MgCl2, 50 mM Tris/HCl pH 7.6, 5 % glycerol, 1 % Triton X-100, 1x complete protease inhibitor). Cleared lysates were incubated with anti-FLAG beads (Sigma A-2220) or with rabbit anti-GFP antibodies (Invitrogen A11222) bound to protein A sepharose (GE Healthcare 17-5280-01) for 2 hours. After three washes with lysis buffer, bound proteins were eluted in SDS sample buffer. Samples were run on 8 % gels and blotted with JL-8 anti-GFP (Clontech, 1:1000), FLAG M2 (Sigma F3165, 1:5000), or myc 9E10 (in house, 1:50) antibodies. Bands were visualized on an Amersham Imager 680.

Quantification and statistical analyses

Phenotypic penetrance was assessed by counting the number of neurons with dendrites still attached to the soma. Here, significance was determined using a categorical two-tailed Fisher’s exact test. Length of unpruned dendrites were measured using the Fiji NeuronJ plugin and compared using the Wilcoxon Mann Whitney test. The number of dendritic kin::lacZ puncta was determined manually and compared using the Wilcoxon Mann Whitney test. EB1 comet orientation was assessed using a categorical two-tailed Fisher’s exact test, and a Mann Whitney test was used to compare comet speed. In order to quantify Shot or EB1 dendrite tip localization, intensity profiles of the distal 10 μm of each dendrite were averaged. The resulting curves were then sectioned into 2 μm segments, and data points from the individual intensity profiles for each segment were pooled and used for statistical comparisons between genotypes using the Mann Whitney test. To compare microtubule density, average cherry-tubulin fluorescence intensities along dendrites were normalized against cherry-tubulin fluorescence in the soma and compared using a Student’s t-test. Presynapse numbers were determined as described (Frommeyer et al, 2026). Two-sample t-tests were performed in Excel, all other tests were performed using Prism 10 software or online (Marx et al, 2016). Where necessary, data were corrected for multiple comparisons using Bennett’s test.